---
title: "SSURGO Integration: Live Queries & Multi-Source Data"
author: "Andrew G. Brown"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    number_sections: true
vignette: >
  %\VignetteIndexEntry{SSURGO Integration}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 6,
  dpi = 100
)
```

# Overview

This vignette demonstrates integration of SSURGO (Soil Survey Geographic) database with geocoda for large-scale soil mapping. Topics covered:

1. **Live SSURGO queries** via soilDB package
2. **Component aggregation** - Weighting within map units
3. **Tile-based workflows** - Spatial optimization for large areas
4. **Integration with field data** - Combining SSURGO with observed samples
5. **Multi-source uncertainty** - Weighting by data quality

## Vignette Status

This is a **minimal scaffold vignette**. It will be filled in with complete examples and demonstrations after Phase 3 and Phase 4 implementation.

## Key Concepts (To Be Expanded)

**SSURGO strengths:**
- Complete spatial coverage (entire USA)
- Standardized component definitions
- Mapunit hierarchies

**Integration approach:**
- Use as spatial prior/context
- Weight by component percentage
- Combine with field observations

**Tile-based optimization:**
- Query large regions efficiently
- Cache results (24-hour TTL)
- Parallel processing

## Placeholder: Complete Example

```{r placeholder, eval=FALSE}
# COMING: Complete SSURGO integration workflow

# 1. Query SSURGO for study area
# ssurgo_data <- soilDB::fetchSDA(...)

# 2. Extract component data
# components <- aggregate_components(ssurgo_data)

# 3. Combine with field observations
# combined <- rbind(field_obs, ssurgo_components)

# 4. Weight by source quality
# weighted <- weight_by_source(combined)

# 5. Model and simulate
# realizations <- gc_sim_composition(model, grid)
```

## References & Resources

- See **Vignette 01** for multi-source data weighting
- See **Vignette 00** for complete workflow
- [soilDB documentation](https://ncss-tech.github.io/soilDB/)

---

**This vignette will be expanded with:**
- Live SDA query examples
- Component aggregation workflows
- Tile-based spatial optimization
- Real-world case studies
- Performance benchmarks
