---
title: "Parameter Selection Guide"
author: "Andrew G. Brown"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Parameter Selection}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 6,
  dpi = 100
)
```

# Overview

Reference guide for selecting variogram and modeling parameters. This vignette provides practical guidance for:

1. **Variogram parameters** - Range, nugget, psill selection
2. **Computational parameters** - Sample size, grid resolution
3. **Hierarchical model tuning** - Shrinkage strength, backend selection
4. **Sensitivity analysis** - How parameters affect predictions

## Vignette Status

**Minimal scaffold** - Expanded from Vignette 10 (Parameter Selection Guide). Will be enhanced with:
- Interactive parameter calculators
- Sensitivity analysis plots
- Real-world case studies
- Decision flowcharts

## Key Topics (To Be Expanded)

### Variogram Range Selection

The range parameter defines spatial correlation decay distance.

**Default:** Domain diagonal / 3

**Scaling guidance:**
- 0.5x: Fine-scale local variation
- 1.0x: Standard (recommended)
- 1.5x: Large-scale smooth trends
- 2.0x: Very smooth fields

### Nugget Parameter

Nugget represents measurement error + local variability.

- Typical: 10-30% of sill
- Small: Smooth surfaces
- Large: Rough surfaces with noise

### Psill Parameter

Partial sill (maximum variogram value - nugget).

- Controls variance amplitude
- Must be positive
- Often normalized to 1.0

## Placeholder: Decision Tree

```{r decision_tree, eval=FALSE}
# COMING: Interactive parameter selection guide

# Question flow:
# 1. What is your domain size?
# 2. How many samples do you have?
# 3. What temporal/spatial resolution is needed?
# 4. How smooth should predictions be?

# Returns: Parameter recommendations with ranges
```

## References

See **Vignette 00** for workflow context.

---

**This vignette will be expanded with:**
- Parameter sensitivity analysis
- Visual parameter effects
- Calculator tools
- Domain-specific guidance (soil texture, geochemistry, etc.)
- Cross-validation optimization workflows
