---
title: "FAQ & Troubleshooting"
author: "Andrew G. Brown"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    number_sections: true
vignette: >
  %\VignetteIndexEntry{FAQ & Troubleshooting}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 6,
  dpi = 100
)
```

# Overview

Frequently asked questions and solutions for common issues when using geocoda for compositional soil mapping.

## Vignette Status

**Minimal scaffold** - Will be expanded with:
- Common error messages and solutions
- Parameter tuning guidelines
- Performance optimization tips
- Data preparation troubleshooting
- Simulation validation checks
- Backend-specific issues (analytical/Stan/Nimble)

## Quick Reference: Common Issues

### Data Preparation

**Q: "Compositional sum not equal to 100%"**
- COMING: Troubleshooting guide for data quality checks

**Q: "Too many zeros in my data"**
- COMING: Zero handling strategies and imputation guidance

**Q: "Warnings about sparse data"**
- COMING: Bootstrap resampling and edge extension techniques

### Variogram Modeling

**Q: "Variogram fitting failed"**
- COMING: Variogram fitting diagnostics and debugging

**Q: "Which range parameter should I use?"**
- See Vignette 06: Parameter Selection Guide

### Simulation & Constraints

**Q: "Simulations violate compositional constraints"**
- COMING: Constraint validation and enforcement troubleshooting

**Q: "Predictions look unrealistic"**
- COMING: Validation checklist and sensitivity analysis

### Hierarchical Models

**Q: "Which backend should I use?"**
- See Vignette 03: Hierarchical Modeling for backend comparison

**Q: "Shrinkage weights don't look right"**
- COMING: Interpreting and tuning shrinkage parameters

**Q: "MCMC chains not converging"**
- COMING: Stan/Nimble convergence diagnostics and debugging

### Risk Assessment

**Q: "How do I interpret probability maps?"**
- See Vignette 02: Ensemble Analysis & Risk Assessment

**Q: "What loss function should I use?"**
- COMING: Loss function selection for different scenarios

## Common Error Messages

```{r error_messages, eval=FALSE}
# COMING: Comprehensive error troubleshooting guide

# Error 1: "object ... not found"
# Solution: ...

# Error 2: "Invalid CRS"
# Solution: ...

# Error 3: "Dimension mismatch"
# Solution: ...
```

## Performance Optimization

```{r performance_tips, eval=FALSE}
# COMING: Performance optimization guidelines

# Tip 1: Reduce grid resolution for exploratory analysis
# Tip 2: Use analytical backend for initial fitting
# Tip 3: Parallelize multiple realizations
# Tip 4: Cache variogram fits
```

## Validation Workflows

```{r validation_checklist, eval=FALSE}
# COMING: Pre-simulation and post-simulation validation checklists

# Pre-simulation checks:
# - Compositional constraint satisfaction
# - Spatial coverage adequacy
# - Outlier detection
# - Stationarity assessment

# Post-simulation checks:
# - Constraint satisfaction in realizations
# - Statistical reproduction (mean, variance, spatial structure)
# - Visual plausibility assessment
# - Comparison with input observations
```

## When to Contact Maintainers

- Unexpected computational errors
- Suspected bugs in constraint enforcement
- Feature requests aligned with roadmap
- Documentation unclear on specific topics

# Integration with Other Vignettes

- See **Vignette 00** for complete workflow overview
- See **Vignette 01** for data preparation guidance
- See **Vignette 02** for risk assessment concepts
- See **Vignette 03** for hierarchical model backends
- See **Vignette 06** for parameter selection

# References

GitHub Issues: https://github.com/brownag/geocoda/issues
Package Documentation: https://brownag.github.io/geocoda

---

**This FAQ will be expanded with:**
- Detailed error troubleshooting
- Performance optimization strategies
- Workflow best practices
- Real-world solutions to common problems
