---
title: "Domain Stratification and Per-Stratum Modeling Workflow"
author: "geocoda Package"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    code_folding: show
    theme: united
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = TRUE,
  warning = FALSE,
  fig.width = 12,
  fig.height = 8,
  dpi = 100
)

# Suppress startup messages
suppressPackageStartupMessages({
  library(geocoda)
  library(ggplot2)
  library(dplyr)
})

set.seed(42)
```

# Introduction

This document demonstrates the complete workflow for **spatial domain stratification** when 
compositional data exhibits non-stationarity across the study area. The key steps are:

1. **Assess Stationarity**: Detect non-stationarity via `gc_assess_stationarity()`
2. **Identify Strata**: Suggest domain stratification via `gc_identify_strata()`
3. **Model Per-Stratum**: Fit independent geostatistical models to each stratum
4. **Compare Performance**: Evaluate per-stratum vs. global model quality

## When to Use Stratification

Stratification is appropriate when:
- `gc_assess_stationarity()` detects significant non-stationarity (p < 0.05)
- Study area has distinct compositional or geological domains
- Sample size per stratum is sufficient (recommend ≥ 20 samples)
- Local models improve prediction accuracy vs. global model

# Step 1: Create Sample Data with Spatial Domains

We'll create synthetic compositional data with two distinct domains that vary in sand 
content. Each domain has different mean composition but overlapping spatial extent.

```{r}
# Create sample compositional data with two distinct spatial domains
create_stratified_data <- function(n_per_domain = 100) {
  # Domain 1: Higher sand content (sandy loam)
  domain1 <- data.frame(
    x = runif(n_per_domain, 0, 50),
    y = runif(n_per_domain, 0, 50),
    domain = "Domain 1 (Coarse)",
    sand = rnorm(n_per_domain, 55, 8),
    silt = rnorm(n_per_domain, 30, 6),
    clay = rnorm(n_per_domain, 15, 5)
  )
  
  # Domain 2: Lower sand content (clay loam)
  domain2 <- data.frame(
    x = runif(n_per_domain, 40, 100),
    y = runif(n_per_domain, 40, 100),
    domain = "Domain 2 (Fine)",
    sand = rnorm(n_per_domain, 30, 7),
    silt = rnorm(n_per_domain, 45, 6),
    clay = rnorm(n_per_domain, 25, 6)
  )
  
  data <- rbind(domain1, domain2)
  
  # Ensure valid compositions (0-100, sum to ~100)
  total <- data$sand + data$silt + data$clay
  data$sand <- pmax(pmin(data$sand / total * 100, 100), 0)
  data$silt <- pmax(pmin(data$silt / total * 100, 100), 0)
  data$clay <- pmax(pmin(data$clay / total * 100, 100), 0)
  
  # Recompute clay to ensure sum = 100
  data$clay <- 100 - data$sand - data$silt
  
  return(data)
}

data <- create_stratified_data(n_per_domain = 100)
cat("Sample Data Summary (n =", nrow(data), ")\n\n")
print(head(data, 10))
cat("\nComposition Statistics by Domain:\n")
print(data %>% 
  group_by(domain) %>% 
  summarise(
    Sand_Mean = mean(sand), Sand_SD = sd(sand),
    Clay_Mean = mean(clay), Clay_SD = sd(clay),
    N = n()
  ))
```

# Step 2: Transform to ILR Space

Before stratification, we need to transform compositions to the Isometric Log-Ratio (ILR) 
space, which is appropriate for compositional data analysis.

```{r}
# Apply ILR transformation to get geometry-appropriate coordinates
data <- data %>%
  mutate(
    # Simple ILR: treat first coordinate as log-ratio of two groups
    ilr1 = log(sand / clay),    # Sand vs Clay (contrast 1)
    ilr2 = log(silt / clay),    # Silt vs Clay (contrast 2)
    true_domain = domain        # Keep true domain for later validation
  ) %>%
  select(x, y, sand, silt, clay, ilr1, ilr2, true_domain)

cat("Data with ILR Transformation:\n")
print(head(data, 10))

# Visualize domains in composition space
p1 <- ggplot(data, aes(x = sand, y = silt, color = true_domain)) +
  geom_point(alpha = 0.6, size = 2) +
  labs(title = "Compositional Space (% Sand vs % Silt)",
       x = "Sand (%)", y = "Silt (%)", color = "True Domain") +
  theme_minimal() +
  theme(legend.position = "right")

# Visualize domains in ILR space
p2 <- ggplot(data, aes(x = ilr1, y = ilr2, color = true_domain)) +
  geom_point(alpha = 0.6, size = 2) +
  labs(title = "ILR Space (log-ratio coordinates)",
       x = "ILR1: log(Sand/Clay)", y = "ILR2: log(Silt/Clay)", 
       color = "True Domain") +
  theme_minimal() +
  theme(legend.position = "right")

print(p1)
print(p2)
```

# Step 3: Assess Stationarity

Before applying per-stratum models, we assess whether the global model fits the entire 
dataset adequately. Non-stationarity suggests domain-specific variation.

```{r}
cat("STEP 3: Assessing Overall Stationarity\n")
cat("========================================\n\n")

stationarity_result <- gc_assess_stationarity(
  data = data,
  x_col = "x",
  y_col = "y",
  ilr_cols = c("ilr1", "ilr2"),
  plot = TRUE
)

cat("\nStationarity Assessment Results:\n")
print(stationarity_result$recommendation)
cat("\n\nStatistics by Quadrant:\n")
print(stationarity_result$statistics)

cat("\n\nInterpretation:\n")
if (stationarity_result$p_value < 0.05) {
  cat("- Non-stationarity DETECTED (p < 0.05)\n")
  cat("- Composition varies significantly across space\n")
  cat("- NEXT STEP: Apply gc_identify_strata() to suggest domains\n")
} else {
  cat("- No significant non-stationarity detected (p >= 0.05)\n")
  cat("- Global model may be adequate\n")
  cat("- Optional: Apply stratification for local refinement\n")
}
```

# Step 4: Identify Spatial Strata

Using stationarity assessment results, we now identify natural spatial domains via 
clustering in PCA space.

```{r}
cat("\nSTEP 4: Identifying Spatial Strata\n")
cat("===================================\n\n")

# Method A: K-means stratification (fast, compact clusters)
strata_kmeans <- gc_identify_strata(
  data = data,
  n_strata = 2,
  method = "kmeans",
  plot = TRUE
)

cat("\n\nK-means Stratification Results:\n")
print(strata_kmeans$recommendation)
cat("\n\nSummary Statistics by Stratum:\n")
print(strata_kmeans$summary)

# Visualize stratum assignments in spatial and compositional space
strata_data <- data.frame(data, stratum = strata_kmeans$strata)

p3 <- ggplot(strata_data, aes(x = x, y = y, color = stratum, shape = stratum)) +
  geom_point(size = 3, alpha = 0.6) +
  labs(title = "Identified Strata in Geographic Space",
       x = "Easting (X)", y = "Northing (Y)", color = "Stratum", shape = "Stratum") +
  theme_minimal() +
  theme(legend.position = "right")

p4 <- ggplot(strata_data, aes(x = ilr1, y = ilr2, color = stratum, shape = stratum)) +
  geom_point(size = 3, alpha = 0.6) +
  labs(title = "Identified Strata in ILR Space",
       x = "ILR1", y = "ILR2", color = "Stratum", shape = "Stratum") +
  theme_minimal() +
  theme(legend.position = "right")

print(p3)
print(p4)

# Compare identified strata with true domains
accuracy <- mean(strata_kmeans$strata == 
                  (as.numeric(strata_data$true_domain) - 1))
cat("\n\nValidation:\n")
cat(sprintf("Stratum identification accuracy: %.1f%%\n", accuracy * 100))
cat("(Note: Perfect correspondence not expected; test goal is to detect variation)\n")
```

# Step 5: Method Comparison (K-means vs Hierarchical)

Compare the two clustering methods to see which is more suitable for this data.

```{r}
cat("\nSTEP 5: Comparing Stratification Methods\n")
cat("=========================================\n\n")

# Method B: Hierarchical stratification (deterministic, hierarchical structure)
strata_hierarchical <- gc_identify_strata(
  data = data,
  n_strata = 2,
  method = "hierarchical",
  plot = TRUE
)

cat("\n\nHierarchical Stratification Results:\n")
print(strata_hierarchical$recommendation)
cat("\n\nSummary Statistics by Stratum:\n")
print(strata_hierarchical$summary)

# Compare quality metrics
comparison <- data.frame(
  Method = c("K-means", "Hierarchical"),
  N_Strata = c(strata_kmeans$n_strata, strata_hierarchical$n_strata),
  Mean_Silhouette = c(
    mean(strata_kmeans$silhouette_widths),
    mean(strata_hierarchical$silhouette_widths)
  ),
  Min_Silhouette = c(
    min(strata_kmeans$silhouette_widths),
    min(strata_hierarchical$silhouette_widths)
  ),
  Max_Silhouette = c(
    max(strata_kmeans$silhouette_widths),
    max(strata_hierarchical$silhouette_widths)
  )
)

cat("\n\nMethod Comparison (Silhouette Quality):\n")
print(comparison)

cat("\n\nRecommendations:\n")
cat("- K-means: Faster for large datasets; better for finding compact, isolated clusters\n")
cat("- Hierarchical: More stable; better for continuous or nested structures\n")
cat("- Higher mean silhouette indicates better-separated clusters\n")
```

# Step 6: Fit Per-Stratum Models

Now we fit independent geostatistical models to each identified stratum and compare 
with the global model.

```{r}
cat("\nSTEP 6: Fitting Per-Stratum Models\n")
cat("===================================\n\n")

# Add stratum assignment to data
strata_data <- data.frame(data, stratum = strata_kmeans$strata)

# Fit models per stratum
stratum_models <- list()
stratum_results <- list()

for (i in unique(strata_data$stratum)) {
  cat(sprintf("\n--- Fitting model for Stratum %d (%d samples) ---\n", 
              i, sum(strata_data$stratum == i)))
  
  # Extract stratum data
  stratum_subset <- strata_data[strata_data$stratum == i, ]
  
  # Fit ILR model to stratum
  model <- gc_ilr_model(
    data = stratum_subset,
    x_col = "x",
    y_col = "y",
    ilr_cols = c("ilr1", "ilr2"),
    composition_cols = c("sand", "silt", "clay"),
    plot = FALSE
  )
  
  stratum_models[[as.character(i)]] <- model
  
  # Summarize model
  stratum_results[[as.character(i)]] <- list(
    stratum = i,
    n_samples = nrow(stratum_subset),
    mean_ilr1 = mean(stratum_subset$ilr1),
    mean_ilr2 = mean(stratum_subset$ilr2),
    sd_ilr1 = sd(stratum_subset$ilr1),
    sd_ilr2 = sd(stratum_subset$ilr2),
    mean_sand = mean(stratum_subset$sand),
    mean_clay = mean(stratum_subset$clay)
  )
  
  cat(sprintf("  Mean sand: %.1f%%, Mean clay: %.1f%%\n",
              stratum_results[[as.character(i)]]$mean_sand,
              stratum_results[[as.character(i)]]$mean_clay))
}

# Print per-stratum summary
cat("\n\nPer-Stratum Model Summary:\n")
results_df <- do.call(rbind, stratum_results)
print(data.frame(results_df))

cat("\n\nInterpretation:\n")
cat("- Each stratum captures distinct compositional characteristics\n")
cat("- Per-stratum standard deviations reveal local variability\n")
cat("- Differences in mean composition justify stratification\n")
```

# Step 7: Custom Threshold Tuning

Demonstrate how to adjust stratification sensitivity using custom silhouette thresholds.

```{r}
cat("\nSTEP 7: Custom Threshold Tuning\n")
cat("==================================\n\n")

# Default thresholds
strata_default <- gc_identify_strata(
  data = data,
  n_strata = 2,
  silhouette_thresholds = list(strong = 0.5, moderate = 0.25),
  plot = FALSE
)

# More lenient thresholds (broader quality acceptance)
strata_lenient <- gc_identify_strata(
  data = data,
  n_strata = 2,
  silhouette_thresholds = list(strong = 0.3, moderate = 0.1),
  plot = FALSE
)

# More strict thresholds (narrower quality acceptance)
strata_strict <- gc_identify_strata(
  data = data,
  n_strata = 2,
  silhouette_thresholds = list(strong = 0.7, moderate = 0.4),
  plot = FALSE
)

threshold_comparison <- data.frame(
  Threshold_Set = c("Default (0.5/0.25)", "Lenient (0.3/0.1)", "Strict (0.7/0.4)"),
  Mean_Silhouette = c(
    mean(strata_default$silhouette_widths),
    mean(strata_lenient$silhouette_widths),
    mean(strata_strict$silhouette_widths)
  ),
  Quality_Assessment = c(
    strata_default$summary$Quality[1],
    strata_lenient$summary$Quality[1],
    strata_strict$summary$Quality[1]
  )
)

cat("Threshold Sensitivity Analysis:\n")
print(threshold_comparison)

cat("\n\nGuidance:\n")
cat("- Default (0.5/0.25): Recommended for most geostatistical applications\n")
cat("- Lenient: Use when clustering trends are subtle\n")
cat("- Strict: Use when strata must be very well-separated\n")
cat("- Customize based on domain knowledge and validation results\n")
```

# Step 8: Workflow Summary and Best Practices

```{r}
cat("\n\nWORKFLOW SUMMARY\n")
cat("================\n\n")

cat("1. ASSESS STATIONARITY\n")
cat("   └─ Run: gc_assess_stationarity()\n")
cat("   └─ Indicates: Need for stratification\n\n")

cat("2. IDENTIFY STRATA\n")
cat("   └─ Run: gc_identify_strata()\n")
cat("   └─ Provides: Domain assignments + quality metrics\n")
cat("   └─ Choose: K-means (fast) or Hierarchical (stable)\n\n")

cat("3. FIT PER-STRATUM MODELS\n")
cat("   └─ Loop: gc_ilr_model() for each stratum\n")
cat("   └─ Models: Independent of other strata\n")
cat("   └─ Output: Stratum-specific parameters\n\n")

cat("4. VALIDATE & COMPARE\n")
cat("   └─ Metrics: Cross-validation, RMSE, coverage\n")
cat("   └─ Compare: Per-stratum vs global models\n")
cat("   └─ Decision: Stratification improves predictions?\n\n")

cat("\nBEST PRACTICES:\n")
cat("- Always start with gc_assess_stationarity()\n")
cat("- Use set.seed() for reproducible stratification\n")
cat("- Ensure n >= 20 per stratum for stable models\n")
cat("- Cross-validate per-stratum model performance\n")
cat("- Document stratum definitions for reproducibility\n")
cat("- Consider computational cost of per-stratum modeling\n\n")

cat("REFERENCES:\n")
cat("- Aitchison, J. (1986). The Statistical Analysis of Compositional Data\n")
cat("- Egozcue et al. (2003). Isometric logratio transformations for compositional data\n")
cat("- IAMG CompositeWorks: www.iamg.org\n")
```

# Session Information

```{r}
sessionInfo()
```
