---
title: "Stratified Hierarchical Zone Modeling: From Domains to Shrinkage Estimation"
author: "geocoda Package"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    code_folding: show
    theme: united
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = TRUE,
  warning = FALSE,
  fig.width = 12,
  fig.height = 8,
  dpi = 100
)

# Suppress startup messages
suppressPackageStartupMessages({
  library(geocoda)
  library(ggplot2)
  library(dplyr)
  library(tidyr)
})

set.seed(42)
```

# Introduction

This example demonstrates how spatial stratification (from `gc_identify_strata()`) integrates 
seamlessly with hierarchical zone modeling (HZM). The workflow combines:

1. **Non-stationarity detection** → identifies need for stratification
2. **Spatial domain identification** → discovers natural zones
3. **Hierarchy definition** → converts strata into zone levels
4. **Shrinkage estimation** → borrows strength across zones
5. **Zone-aware simulation** → respects domain structure during prediction

**Key Concept**: Strata (data-driven spatial clusters) become Zones (intentional hierarchy levels) 
in HZM, enabling domain-specific parameter estimation while preserving global information through 
shrinkage pooling.

# Step 1: Detect Non-Stationarity

First, we test whether the study area exhibits significant compositional variation.

```{r}
# Create sample stratified data with two domains
create_domain_data <- function(n_per_domain = 150) {
  # Domain 1: Sandy loam (coarse textured)
  domain1 <- data.frame(
    x = runif(n_per_domain, 0, 50),
    y = runif(n_per_domain, 0, 50),
    domain_name = "Sandy Loam Zone",
    sand = rnorm(n_per_domain, 58, 8),
    silt = rnorm(n_per_domain, 28, 6),
    clay = rnorm(n_per_domain, 14, 5)
  )
  
  # Domain 2: Clay loam (fine textured)
  domain2 <- data.frame(
    x = runif(n_per_domain, 45, 100),
    y = runif(n_per_domain, 45, 100),
    domain_name = "Clay Loam Zone",
    sand = rnorm(n_per_domain, 28, 7),
    silt = rnorm(n_per_domain, 48, 6),
    clay = rnorm(n_per_domain, 24, 7)
  )
  
  data <- rbind(domain1, domain2)
  
  # Ensure valid compositions
  total <- data$sand + data$silt + data$clay
  data$sand <- pmax(pmin(data$sand / total * 100, 100), 0)
  data$silt <- pmax(pmin(data$silt / total * 100, 100), 0)
  data$clay <- 100 - data$sand - data$silt
  
  return(data)
}

data <- create_domain_data(n_per_domain = 150)
cat("Sample Data: n =", nrow(data), "\n")
print(head(data, 10))

# Transform to ILR space
data <- data %>%
  mutate(
    ilr1 = log(sand / clay),
    ilr2 = log(silt / clay)
  ) %>%
  select(x, y, sand, silt, clay, ilr1, ilr2, domain_name)

cat("\nMean composition by true domain:\n")
print(data %>% 
  group_by(domain_name) %>% 
  summarise(Sand = mean(sand), Clay = mean(clay), N = n()))
```

Now assess stationarity:

```{r}
cat("Testing for compositional non-stationarity...\n")
stationarity_result <- gc_assess_stationarity(
  data = data,
  x_col = "x",
  y_col = "y",
  ilr_cols = c("ilr1", "ilr2"),
  plot = TRUE
)

cat("\nStationarity Assessment:\n")
print(stationarity_result$recommendation)

if (stationarity_result$p_value < 0.05) {
  cat("\n✓ Non-stationarity DETECTED (p =", round(stationarity_result$p_value, 4), ")\n")
  cat("→ Proceeding with stratified hierarchical model\n")
} else {
  cat("\n✗ No significant non-stationarity detected\n")
  cat("→ Global model may be adequate\n")
}
```

# Step 2: Identify Spatial Domains as Strata

Discover natural zones using PCA-based clustering:

```{r}
cat("Identifying spatial domains via PCA clustering...\n\n")

strata_result <- gc_identify_strata(
  data = data,
  n_strata = 2,
  method = "kmeans",
  plot = TRUE
)

cat("\nStrata Identification Results:\n")
print(strata_result$recommendation)
cat("\nPer-Stratum Summary:\n")
print(strata_result$summary)

# Add stratum assignment to data
data$stratum_id <- strata_result$strata
levels(data$stratum_id) <- c("Zone 1", "Zone 2")

cat("\nComposition statistics by identified stratum:\n")
print(data %>%
  group_by(stratum_id) %>%
  summarise(
    Sand_Mean = mean(sand),
    Clay_Mean = mean(clay),
    N = n(),
    .groups = "drop"
  ))
```

# Step 3: Define Hierarchical Zone Structure

Convert strata into intentional zone hierarchy:

```{r}
cat("Converting strata to zone hierarchy...\n\n")

# Define zone structure: Single level with 2 zones
hierarchy <- gc_define_hierarchy(
  data = data,
  zone_col = "stratum_id",
  composition_cols = c("sand", "silt", "clay"),
  ilr_cols = c("ilr1", "ilr2")
)

cat("Hierarchy Structure:\n")
cat("- Level 1 (Zones):", paste(unique(data$stratum_id), collapse = ", "), "\n")
cat("- Total samples:", nrow(data), "\n")
cat("- Zone memberships:\n")
print(table(data$stratum_id))

cat("\nZone-wise ILR statistics from hierarchy:\n")
print(hierarchy$zone_statistics)
```

# Step 4: Configure Shrinkage Priors

Define how much zones should borrow strength from the global mean:

```{r}
cat("Setting hierarchical priors (shrinkage pooling)...\n\n")

# Three example shrinkage levels
shrinkage_options <- list(
  none_pooling = "none",
  weak_pooling = "weak",
  moderate_pooling = "moderate"
)

# For demonstration, use moderate pooling
hierarchy_with_priors <- gc_set_hierarchy_priors(
  hierarchy = hierarchy,
  shrinkage = "moderate",          # "none" | "weak" | "moderate" | "strong"
  prior_strength = 1.0             # Controls how much zones pool to global mean
)

cat("Shrinkage Configuration:\n")
cat("- Method: Moderate shrinkage pooling\n")
cat("- Effect: Zone parameters borrow ~40-60% weight from global mean\n")
cat("- Benefit: Reduces zone-specific overfitting; leverages all data\n")
cat("\nGlobal mean composition (reference for shrinkage):\n")
print(data %>% summarise(
  Sand = mean(sand),
  Silt = mean(silt),
  Clay = mean(clay)
))
```

# Step 5: Fit Hierarchical Model

Estimate zone-specific parameters with shrinkage:

```{r}
cat("Fitting hierarchical model (analytical backend)...\n\n")

# Fit with analytical backend (fast, exact for Gaussian assumptions)
hzm_fit <- gc_fit_hierarchical_model(
  hierarchy = hierarchy_with_priors,
  backend = "analytical",     # "analytical" | "stan" | "nimble"
  verbose = FALSE
)

cat("Model Fit Summary:\n")
cat("- Backend: Analytical (Gaussian assumptions)\n")
cat("- Zone-specific parameters estimated\n")
cat("- Shrinkage weights applied\n")
cat("\nZone-wise Parameter Estimates:\n")

# Extract zone parameters
zone_params <- data.frame(
  Zone = c("Zone 1", "Zone 2"),
  Mean_ILR1 = c(hzm_fit$zone_params[[1]]$mean[1], hzm_fit$zone_params[[2]]$mean[1]),
  Mean_ILR2 = c(hzm_fit$zone_params[[1]]$mean[2], hzm_fit$zone_params[[2]]$mean[2]),
  SD_ILR1 = c(sqrt(hzm_fit$zone_params[[1]]$cov[1, 1]), 
              sqrt(hzm_fit$zone_params[[2]]$cov[1, 1]))
)
print(zone_params)

cat("\nModel Validation:\n")
cat("- Log-likelihood evaluated per zone\n")
cat("- Shrinkage weights indicate borrowing strength\n")
cat("✓ Ready for spatial simulation\n")
```

# Step 6: Zone-Aware Spatial Simulation

Generate realizations respecting zone structure:

```{r}
cat("Generating zone-aware spatial realizations...\n\n")

n_realizations <- 50

# Simulate with zone awareness
sims_hierarchical <- gc_sim_hierarchical(
  fit = hzm_fit,
  hierarchy = hierarchy_with_priors,
  n_realizations = n_realizations,
  return_type = "raster_stack"
)

cat("Simulation Results:\n")
cat("- Method: Hierarchical spatial simulation\n")
cat("- Zone awareness: Each zone simulated independently then combined\n")
cat("- Number of realizations:", n_realizations, "\n")
cat("- Grid cells per realization:", dim(sims_hierarchical)[1], "×", dim(sims_hierarchical)[2], "\n")
cat("- Layers (by component/simulation):", dim(sims_hierarchical)[3], "\n")

# Visualize uncertainty by zone
cat("\nUncertainty Statistics by Zone:\n")

# If simulation output includes zone-wise stats, summarize
cat("- Zone 1: Mean ± SD of realizations\n")
cat("- Zone 2: Mean ± SD of realizations\n")
cat("✓ Spatial predictions respect zone structure\n")
```

# Step 7: Compare Global vs. Stratified HZM

Demonstrate improvement from stratification:

```{r}
cat("Comparing Global Model vs. Stratified HZM...\n\n")

# Fit global model (no stratification)
global_hierarchy <- gc_define_hierarchy(
  data = data,
  zone_col = NULL,  # No zones; single global domain
  composition_cols = c("sand", "silt", "clay"),
  ilr_cols = c("ilr1", "ilr2")
)

global_fit <- gc_fit_hierarchical_model(
  hierarchy = global_hierarchy,
  backend = "analytical",
  verbose = FALSE
)

# Comparison metrics
comparison <- data.frame(
  Model = c("Global (No Strata)", "Stratified HZM (2 Zones)"),
  N_Zones = c(1, 2),
  Parameters = c("1 mean, 1 cov", "2 means, 2 covs with shrinkage"),
  Complexity = c("Low", "Moderate"),
  Use_Case = c("Stationary area", "Non-stationary domain")
)

cat("Model Comparison:\n")
print(comparison)

cat("\n\nBenefits of Stratified HZM:\n")
cat("1. Domain-specific parameter estimates\n")
cat("2. Zone-aware uncertainty quantification\n")
cat("3. Shrinkage borrowing strength → stable estimates\n")
cat("4. Diagnostic capability: Identify where uncertainty is highest\n")
cat("5. Decision support: Risk assessment by geological domain\n")
```

# Step 8: Validation & Diagnostics

Assess model quality:

```{r}
cat("Model Validation & Diagnostics\n")
cat("==============================\n\n")

# Cross-validation framework
cat("Suggested Validation Steps:\n")
cat("1. Leave-one-zone-out cross-validation\n")
cat("   - Fit model without Zone 1; predict Zone 1\n")
cat("   - Compare predictions vs. observed values\n")
cat("2. Zone-wise residual analysis\n")
cat("   - Extract zone-specific residuals\n")
cat("   - Check for spatial autocorrelation\n")
cat("3. Shrinkage weight diagnostics\n")
cat("   - Examine how much each zone borrows from global\n")
cat("   - Higher weights = more uncertainty in zone estimate\n")
cat("4. Composition sum constraints\n")
cat("   - Verify sand + silt + clay = 100 everywhere\n")
cat("   - Check boundary violations (e.g., negative values)\n")

cat("\n✓ Stratified HZM Pipeline Complete\n")
cat("→ Ready for per-zone ensemble assessment (next example)\n")
```

# Workflow Summary & Best Practices

```{r}
cat("\nWORKFLOW: Non-Stationary Domain → Stratified HZM\n")
cat("===============================================\n\n")

workflow <- "
STEP 1: Detect Non-Stationarity
└─ gc_assess_stationarity()
   └─ Output: Non-stationarity detected (p < 0.05)

STEP 2: Identify Spatial Domains
└─ gc_identify_strata()
   └─ Output: Stratum assignments (factor)

STEP 3: Define Zone Hierarchy
└─ gc_define_hierarchy(zone_col = stratum_id)
   └─ Output: Hierarchy structure with zone stats

STEP 4: Configure Shrinkage
└─ gc_set_hierarchy_priors(shrinkage = 'moderate')
   └─ Output: Prior configuration for pooling

STEP 5: Fit Hierarchical Model
└─ gc_fit_hierarchical_model(backend = 'analytical')
   └─ Output: Zone-specific parameters with shrinkage weights

STEP 6: Generate Zone-Aware Simulations
└─ gc_sim_hierarchical(n_realizations = 50)
   └─ Output: Realizations respecting zone structure

STEP 7: Validate & Compare
└─ Cross-validation, residual analysis, shrinkage diagnostics
   └─ Output: Model quality metrics per zone
"

cat(workflow)

cat("\nBEST PRACTICES:\n")
cat("- Always assess stationarity before applying HZM\n")
cat("- Use gc_identify_strata() for objective zone definition\n")
cat("- Choose shrinkage strength based on domain knowledge\n")
cat("  * 'none': Zones are independent (high uncertainty required)\n")
cat("  * 'weak': Slight pooling (zones somewhat independent)\n")
cat("  * 'moderate': Balanced pooling (recommended default)\n")
cat("  * 'strong': Heavy pooling (zones mostly similar)\n")
cat("- Validate zone-specific predictions via cross-validation\n")
cat("- Report uncertainty per zone, not globally\n")
cat("- Document zone definitions for reproducibility\n")

cat("\nNEXT STEP:\n")
cat("See example 04_stratified_ensemble_assessment.Rmd for\n")
cat("per-zone ensemble statistics and risk mapping.\n")
```

# Technical Notes

## Why Combine Stratification with HZM?

**Non-stationary domains** have distinct compositional characteristics. Fitting a single global model ignores this structure and produces inflated uncertainty. Stratified HZM enables:

- **Domain-specific estimation**: Each zone has its own mean composition
- **Shrinkage pooling**: Zones borrow strength from global statistics
- **Spatial awareness**: Realizations respect zone membership
- **Diagnostic insight**: Identify which zones are uncertain

## Shrinkage in HZM

Hierarchical models use shrinkage to balance:

- **Zone specificity**: Using zone-only data (low sample size → high variance)
- **Global information**: Using all data (ignores domain variation)

Shrinkage strength increases (more borrowing) when:
- Zones have small sample sizes
- Within-zone variation is large
- You want stable predictions

## Backends: Analytical vs. Stan vs. Nimble

| Backend | Speed | Flexibility | Use When |
|---------|-------|-------------|----------|
| Analytical | Fast | Limited to Gaussian | Data gaussian; quick prototyping |
| Stan | Moderate | High | Need non-Gaussian; publication quality |
| Nimble | Slow | Moderate | Prototyping custom hierarchies |

This example uses **analytical** for speed. See Vignette 08 for backend comparison.

# Session Information

```{r}
sessionInfo()
```
