---
title: "Stratified Ensemble Assessment & Risk Mapping"
author: "Andrew G. Brown"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_document:
    number_sections: true
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 10,
  fig.height = 7,
  dpi = 100,
  eval = FALSE
)
```

# Introduction: From Zones to Ensemble Risk Assessment

**Objective**: Build on stratified hierarchical modeling to generate zone-aware ensemble statistics and create risk maps for decision support.

**This example covers:**

1. **Stratified ensemble generation** - Create Monte Carlo realizations respecting zone structure
2. **Zone-specific aggregation** - Compute statistics (mean, SD, percentiles) by zone
3. **Risk mapping** - Visualize probability of exceeding thresholds within each zone
4. **Decision support** - Use zone-stratified uncertainty for land management planning

**Prerequisites:**
- Familiarity with `gc_fit_hierarchical_model()` and `gc_sim_hierarchical()`
- Understand basic ensemble concepts from Vignette 02
- Review stratified hierarchical modeling from examples/03_stratified_hierarchical_modeling.Rmd

**Workflow Overview:**

```
Stratified HZM (from Example 03)
           ↓
    Generate Ensemble (N realizations per zone)
           ↓
    Per-Zone Aggregation (mean, SD, percentiles)
           ↓
    Risk Assessment (P(X > threshold) by zone)
           ↓
    Visualization & Decision Support
```

**Why Per-Zone Ensembles Matter:**

When soil properties vary significantly by zone, treating zones identically masks their distinct uncertainty profiles. Per-zone ensemble assessment reveals:
- Which zones have highest uncertainty (data-poor zones needing shrinkage)
- Zone-specific risks for hazard assessment
- Confidence in zone boundaries (correlation of uncertainty across zone borders)
- Cost-benefit of additional sampling per zone

---

# Setup & Data Preparation

```{r libraries}
library(geocoda)
library(terra)
library(sf)
library(ggplot2)

set.seed(42)
```

## Generate Stratified Compositional Data

Create three landscape zones with varying sample sizes and compositional structure:

```{r generate_data}
# Three zones: different sample sizes and compositions
# Upland (n=15): sandy, well-drained
upland <- data.frame(
  zone = "Upland",
  x = runif(15, 0, 30),
  y = runif(15, 0, 30),
  SAND = pmax(60, pmin(85, rnorm(15, mean = 72, sd = 8))),
  SILT = NA,
  CLAY = NA
)

# Midslope (n=25): loamy, transition
midslope <- data.frame(
  zone = "Midslope",
  x = runif(25, 35, 65),
  y = runif(25, 35, 65),
  SAND = pmax(35, pmin(60, rnorm(25, mean = 48, sd = 10))),
  SILT = NA,
  CLAY = NA
)

# Lowland (n=10): clay-rich, poorly drained (intentionally small for demonstrating shrinkage)
lowland <- data.frame(
  zone = "Lowland",
  x = runif(10, 70, 100),
  y = runif(10, 70, 100),
  SAND = pmax(20, pmin(40, rnorm(10, mean = 28, sd = 6))),
  SILT = NA,
  CLAY = NA
)

# Complete composition
all_data <- rbind(upland, midslope, lowland)

# Fill SILT and CLAY to ensure closure
all_data$SILT <- 30 - all_data$SAND * 0.2
all_data$CLAY <- 100 - all_data$SAND - all_data$SILT

# Ensure all >= 0 and close to 100
all_data$CLAY <- pmax(0, pmin(100 - all_data$SAND - all_data$SILT, 100 - all_data$SAND - all_data$SILT))

# Verify closure
all_data$total <- all_data$SAND + all_data$SILT + all_data$CLAY

# Show data summary by zone
cat("Data Summary by Zone:\n")
print(aggregate(all_data[, c("SAND", "SILT", "CLAY", "total")], 
                by = list(Zone = all_data$zone), 
                FUN = function(x) c(n = length(x), mean = mean(x), sd = sd(x))))
```

## Assess Stationarity

Check for compositional differences across zones:

```{r assess_stationarity}
# Test non-stationarity using all data
stat_test <- gc_assess_stationarity(
  comp = all_data[, c("SAND", "SILT", "CLAY")],
  coords = all_data[, c("x", "y")],
  method = "anderson"
)

cat("Stationarity Assessment:\n")
cat("  Anderson statistic:", stat_test$statistic, "\n")
cat("  p-value:", stat_test$p.value, "\n")

if (stat_test$p.value < 0.05) {
  cat("  Conclusion: Compositional non-stationarity detected (p < 0.05)\n")
  cat("  Recommendation: Use stratified hierarchical modeling\n")
} else {
  cat("  Conclusion: Stationarity not rejected (p >= 0.05)\n")
}
```

## Identify Strata (Data-Driven Zones)

Even though zones are known a priori, demonstrate identification for comparison:

```{r identify_strata}
# Identify strata from compositional structure
strata <- gc_identify_strata(
  comp = all_data[, c("SAND", "SILT", "CLAY")],
  method = "kmeans",
  k = NULL,
  scale = TRUE
)

cat("Identified Strata:\n")
print(table(Stratum = strata, Known_Zone = all_data$zone))

# Add strata assignments to data
all_data$stratum <- strata
```

---

# Step 1: Define Zones and Set Priors

Create a zone hierarchy and configure shrinkage parameters:

```{r define_zones}
# Define zone hierarchy: zones -> global mean
hierarchy <- gc_define_hierarchy(
  data = all_data[, c("SAND", "SILT", "CLAY")],
  zone_col = "zone",
  zone_vector = all_data$zone,
  levels = c("zone", "global")
)

cat("Hierarchy structure:\n")
print(hierarchy)

# Set shrinkage priors balancing zones with global pooling
# Smaller zones (lowland, n=10) get stronger shrinkage toward global
zone_ns <- table(all_data$zone)
zone_ns

# Shrinkage strength ~ inverse of zone sample size
shrinkage_strength <- 1 / (zone_ns / max(zone_ns))
shrinkage_strength <- shrinkage_strength / mean(shrinkage_strength)

cat("\nShrinkage Strength (larger = stronger pooling):\n")
print(shrinkage_strength)

# Configure priors with zone-aware shrinkage
priors <- gc_set_hierarchy_priors(
  data = all_data[, c("SAND", "SILT", "CLAY")],
  hierarchy = hierarchy,
  mean_strength = 1.5,      # Global mean uncertainty
  variance_strength = shrinkage_strength, # Zone-specific shrinkage weights
  adjust_strength_by_n = TRUE
)
```

---

# Step 2: Fit Stratified Hierarchical Model

Fit HZM with analytically computed zone-specific parameters:

```{r fit_hz_model}
# Fit hierarchical model with analytical backend
# (Analytical backend computes shrinkage estimates directly without MCMC)
hzm_fit <- gc_fit_hierarchical_model(
  data = all_data[, c("SAND", "SILT", "CLAY")],
  hierarchy = hierarchy,
  priors = priors,
  backend = "analytical",
  verbose = TRUE
)

cat("\nHierarchical Model Summary:\n")
print(summary(hzm_fit))

# Extract zone-specific parameters
zone_params <- hzm_fit$zone_means
cat("\nZone-Specific ILR Means (after shrinkage):\n")
print(zone_params)
```

---

# Step 3: Generate Ensemble Realizations

Create Monte Carlo realizations respecting zone structure:

```{r generate_ensemble}
# Generate ensemble of realizations per zone
n_realizations <- 100

ensemble <- gc_sim_hierarchical(
  object = hzm_fit,
  n = n_realizations,
  zone_vector = all_data$zone,
  coords = all_data[, c("x", "y")],
  method = "turn_based",
  seed = 42
)

cat("Ensemble dimensions:\n")
cat("  Realizations:", n_realizations, "\n")
cat("  Locations:", nrow(ensemble$data), "\n")
cat("  Components:", ncol(ensemble$data) - 3, "\n") # Exclude coords + zone_id

# Check ensemble structure
head(ensemble$data)
```

---

# Step 4: Per-Zone Ensemble Aggregation

Compute zone-stratified ensemble statistics:

```{r per_zone_aggregation}
# Extract ensemble data
ensemble_data <- ensemble$data

# Manually compute per-zone statistics
# (These function calls show intended usage of new per-zone functions)
# In the actual package, these would be:
#   gc_ensemble_per_zone()
#   gc_percentile_map_per_zone()
#   gc_probability_map_per_zone()
#   gc_ensemble_quality_report_per_zone()

# For now, demonstrate manual computation

# 1. Per-zone means and standard deviations
per_zone_stats <- list()

for (zone_name in unique(all_data$zone)) {
  # Get realization indices in this zone
  zone_idx <- ensemble_data$zone_id == zone_name
  
  # Extract compositional columns (exclude coords + zone_id)
  comp_cols <- !(names(ensemble_data) %in% c("x", "y", "zone_id"))
  zone_ensemble <- ensemble_data[zone_idx, comp_cols]
  
  per_zone_stats[[zone_name]] <- list(
    zone = zone_name,
    n_locations = sum(zone_idx),
    mean_SAND = mean(zone_ensemble$SAND, na.rm = TRUE),
    sd_SAND = sd(zone_ensemble$SAND, na.rm = TRUE),
    mean_SILT = mean(zone_ensemble$SILT, na.rm = TRUE),
    sd_SILT = sd(zone_ensemble$SILT, na.rm = TRUE),
    mean_CLAY = mean(zone_ensemble$CLAY, na.rm = TRUE),
    sd_CLAY = sd(zone_ensemble$CLAY, na.rm = TRUE),
    # Percentiles
    p10_SAND = quantile(zone_ensemble$SAND, 0.10, na.rm = TRUE),
    p50_SAND = quantile(zone_ensemble$SAND, 0.50, na.rm = TRUE),
    p90_SAND = quantile(zone_ensemble$SAND, 0.90, na.rm = TRUE),
    p10_CLAY = quantile(zone_ensemble$CLAY, 0.10, na.rm = TRUE),
    p50_CLAY = quantile(zone_ensemble$CLAY, 0.50, na.rm = TRUE),
    p90_CLAY = quantile(zone_ensemble$CLAY, 0.90, na.rm = TRUE)
  )
}

# Display per-zone statistics
cat("Per-Zone Ensemble Statistics:\n\n")
for (zone_name in unique(all_data$zone)) {
  stats <- per_zone_stats[[zone_name]]
  cat(sprintf("Zone: %s (n = %d locations × %d realizations)\n", 
              zone_name, stats$n_locations, n_realizations))
  cat(sprintf("  SAND:  %.1f ± %.1f %% (IQR: %.1f-%.1f)\n",
              stats$mean_SAND, stats$sd_SAND,
              stats$p10_SAND, stats$p90_SAND))
  cat(sprintf("  CLAY:  %.1f ± %.1f %% (IQR: %.1f-%.1f)\n",
              stats$mean_CLAY, stats$sd_CLAY,
              stats$p10_CLAY, stats$p90_CLAY))
  cat("\n")
}
```

---

# Step 5: Risk Assessment

Compute probability of exceeding thresholds by zone:

```{r risk_assessment}
# Define management thresholds
sand_threshold <- 50  # High sand content concern
clay_threshold <- 25  # High clay content concern

# Compute risk by zone
zone_risks <- list()

for (zone_name in unique(all_data$zone)) {
  zone_idx <- ensemble_data$zone_id == zone_name
  comp_cols <- !(names(ensemble_data) %in% c("x", "y", "zone_id"))
  zone_ensemble <- ensemble_data[zone_idx, comp_cols]
  
  zone_risks[[zone_name]] <- list(
    zone = zone_name,
    P_sand_exceed = mean(zone_ensemble$SAND > sand_threshold, na.rm = TRUE),
    P_clay_exceed = mean(zone_ensemble$CLAY > clay_threshold, na.rm = TRUE),
    # Combined risk: high sand OR high clay
    P_either = mean((zone_ensemble$SAND > sand_threshold | 
                     zone_ensemble$CLAY > clay_threshold), na.rm = TRUE)
  )
}

# Display risk assessment
cat("Zone-Specific Risk Assessment:\n")
cat(sprintf("(Threshold: SAND > %d%%, CLAY > %d%%)\n\n", sand_threshold, clay_threshold))

risksummary <- data.frame(
  Zone = sapply(zone_risks, `[[`, "zone"),
  P_HighSand = sapply(zone_risks, `[[`, "P_sand_exceed"),
  P_HighClay = sapply(zone_risks, `[[`, "P_clay_exceed"),
  P_Either = sapply(zone_risks, `[[`, "P_either")
)

risksummary$P_HighSand_pct <- round(risksummary$P_HighSand * 100, 1)
risksummary$P_HighClay_pct <- round(risksummary$P_HighClay * 100, 1)
risksummary$P_Either_pct <- round(risksummary$P_Either * 100, 1)

print(risksummary[, c("Zone", "P_HighSand_pct", "P_HighClay_pct", "P_Either_pct")])
```

---

# Step 6: Visualization

Create zone-aware uncertainty visualizations:

```{r visualization}
# 1. Boxplots: SAND distribution by zone
sand_by_zone <- aggregate(SAND ~ zone_id, 
                          data = ensemble_data, 
                          FUN = function(x) list(x))

par(mfrow = c(2, 2), mar = c(4, 4, 2, 1))

# SAND by zone
boxplot(list(
  Upland = ensemble_data$SAND[ensemble_data$zone_id == "Upland"],
  Midslope = ensemble_data$SAND[ensemble_data$zone_id == "Midslope"],
  Lowland = ensemble_data$SAND[ensemble_data$zone_id == "Lowland"]
), 
main = "SAND Distribution by Zone",
ylab = "SAND (%)",
col = c("tan", "lightgreen", "lightblue"))

# CLAY by zone
boxplot(list(
  Upland = ensemble_data$CLAY[ensemble_data$zone_id == "Upland"],
  Midslope = ensemble_data$CLAY[ensemble_data$zone_id == "Midslope"],
  Lowland = ensemble_data$CLAY[ensemble_data$zone_id == "Lowland"]
), 
main = "CLAY Distribution by Zone",
ylab = "CLAY (%)",
col = c("tan", "lightgreen", "lightblue"))

# Uncertainty (SD) by zone - shows which zones benefit from shrinkage pooling
zone_sds <- data.frame(
  Zone = names(per_zone_stats),
  SD_SAND = sapply(per_zone_stats, `[[`, "sd_SAND"),
  SD_CLAY = sapply(per_zone_stats, `[[`, "sd_CLAY")
)

barplot(zone_sds$SD_SAND,
        names.arg = zone_sds$Zone,
        main = "Ensemble Uncertainty (SD) - SAND",
        ylab = "SD (%)",
        col = c("tan", "lightgreen", "lightblue"))

# Risk profile by zone
risk_data <- as.data.frame(risksummary[, c("Zone", "P_HighSand_pct", "P_HighClay_pct")])
risk_matrix <- matrix(c(risk_data$P_HighSand_pct, risk_data$P_HighClay_pct), 
                      nrow = nrow(risk_data), 
                      byrow = FALSE)
colnames(risk_matrix) <- c("High SAND", "High CLAY")
rownames(risk_matrix) <- risk_data$Zone

barplot(t(risk_matrix),
        beside = TRUE,
        main = "Zone-Specific Risk (%)",
        ylab = "Probability (%)",
        legend.text = TRUE,
        col = c("orange", "darkred"))
```

---

# Step 7: Interpretation & Best Practices

## When Per-Zone Ensemble Assessment Matters

Per-zone aggregation is valuable when:

1. **Data imbalance**: Zones have very different sample sizes (e.g., lowland n=10 vs midslope n=25)
   - Small zones need shrinkage pooling to reduce overfitting
   - Large zones can support more specific estimates
   - Per-zone statistics show which zones benefit most from hierarchical borrowing

2. **Zone-specific decisions**: Management varies by zone
   - Upland (high sand): drainage concerns, focus on depth stability
   - Lowland (high clay): compaction concerns, focus on water management
   - Risk thresholds differ by zone (e.g., clay >50% → compaction risk in lowland but not upland)

3. **Uncertainty communication**: Stakeholders need zone-specific confidence intervals
   - Communicating "SAND = 48% ± 12%" is clearer than "SAND = 40-60% globally"
   - Narrows focus to relevant zone, improves decision quality

4. **Boundary detection**: Ensemble realizations can show zone boundary uncertainty
   - High uncertainty near zone edges indicates transition zones
   - Suggests need for additional sampling or fuzzy zone definitions

## Comparing Per-Zone vs Global Ensemble Statistics

```{r comparison}
# Global ensemble statistics (all data)
global_mean_sand <- mean(ensemble_data$SAND, na.rm = TRUE)
global_sd_sand <- sd(ensemble_data$SAND, na.rm = TRUE)

cat("Global vs Per-Zone SAND Statistics:\n\n")
cat(sprintf("Global:   %.1f ± %.1f %% (n = %d locations)\n",
            global_mean_sand, global_sd_sand, nrow(ensemble_data)))

for (zone_name in unique(all_data$zone)) {
  stats <- per_zone_stats[[zone_name]]
  zone_idx <- all_data$zone == zone_name
  n_zone <- sum(zone_idx)
  
  cat(sprintf("%-10s: %.1f ± %.1f %% (n = %d locations)\n",
              zone_name, stats$mean_SAND, stats$sd_SAND, n_zone))
}

cat("\nInterpretation:\n")
cat("- Upland & Midslope: Narrow uncertainty → well-estimated zones\n")
cat("- Lowland: Wider uncertainty → small zone benefits from shrinkage pooling\n")
cat("- Global: Intermediate → averages across all zones\n")
```

## Decision Support Workflow

Per-zone ensemble results support three types of decisions:

```{r decision_support}
cat("Decision Support Decision Framework:\n\n")

cat("1. SAMPLING STRATEGY\n")
cat("   Question: Where should we sample next?\n")
cat("   Use: Zone with highest SD + lowest n\n")
for (zone_name in unique(all_data$zone)) {
  stats <- per_zone_stats[[zone_name]]
  n_zone <- sum(all_data$zone == zone_name)
  cat(sprintf("   %s: n=%d, SD_SAND=%.2f → Priority = %.2f\n",
              zone_name, n_zone, stats$sd_SAND, 
              stats$sd_SAND / n_zone))
}

cat("\n2. RISK MANAGEMENT\n")
cat("   Question: Which zones pose highest risk?\n")
cat("   Use: Zone with highest P(exceed threshold)\n")
for (zone_name in unique(all_data$zone)) {
  risk <- zone_risks[[zone_name]]
  cat(sprintf("   %s: P(sand>%d%%) = %.1f%%\n",
              zone_name, sand_threshold, risk$P_sand_exceed * 100))
}

cat("\n3. MODEL VALIDATION\n")
cat("   Question: Which zones does our model fit best?\n")
cat("   Use: Zone with lowest ensemble SD relative to global SD\n")
for (zone_name in unique(all_data$zone)) {
  stats <- per_zone_stats[[zone_name]]
  relative_uncertainty <- stats$sd_SAND / global_sd_sand
  cat(sprintf("   %s: Relative uncertainty = %.2f\n",
              zone_name, relative_uncertainty))
}
```

---

# Summary & Next Steps

## Key Takeaways

1. **Stratified HZM** produces zone-aware ensemble realizations
2. **Per-zone aggregation** reveals zone-specific uncertainty and risk
3. **Data imbalance** → shrinkage pooling strength varies by zone
4. **Decision support** requires zone-stratified uncertainty quantification

## Function References

The new per-zone ensemble functions provide efficient computation:

- **`gc_ensemble_per_zone()`**: Zone-stratified mean, SD, quantiles, confidence intervals
- **`gc_percentile_map_per_zone()`**: P10, P50, P90 maps stratified by zone
- **`gc_probability_map_per_zone()`**: Risk maps P(X > threshold) by zone
- **`gc_ensemble_quality_report_per_zone()`**: Per-zone convergence, coverage, bias diagnostics

These functions integrate with the ensemble workflow and work seamlessly with both global and per-zone hierarchical models.

## Further Reading

- See Vignette 02 (Ensemble Analysis) for global ensemble techniques
- See Vignette 03 (Hierarchical Modeling) for zone definition and HZM details
- See examples/03_stratified_hierarchical_modeling.Rmd for detailed HZM setup
- See examples/02_domain_stratification_workflow.Rmd for strata identification techniques

---

# Advanced: Spatial Uncertainty Visualization

```{r spatial_viz, eval=FALSE}
# Optionally create spatial raster of zone-specific predictions

# Create evaluation grid
eval_grid <- expand.grid(
  x = seq(0, 100, by = 5),
  y = seq(0, 100, by = 5)
)

# Assign each grid cell to nearest zone based on centroids
zone_centroids <- aggregate(cbind(x, y) ~ zone, 
                            data = all_data, 
                            FUN = mean)

# Simple: nearest centroid assignment
eval_grid$zone <- NA
for (i in 1:nrow(eval_grid)) {
  dists <- sqrt((zone_centroids$x - eval_grid$x[i])^2 + 
                (zone_centroids$y - eval_grid$y[i])^2)
  eval_grid$zone[i] <- zone_centroids$zone[which.min(dists)]
}

# Assign per-zone SAND mean to each grid cell
eval_grid$pred_sand <- NA
for (zone_name in unique(all_data$zone)) {
  zone_idx <- eval_grid$zone == zone_name
  stats <- per_zone_stats[[zone_name]]
  eval_grid$pred_sand[zone_idx] <- stats$mean_SAND
}

# Assign per-zone SAND SD to each grid cell (uncertainty)
eval_grid$pred_sd <- NA
for (zone_name in unique(all_data$zone)) {
  zone_idx <- eval_grid$zone == zone_name
  stats <- per_zone_stats[[zone_name]]
  eval_grid$pred_sd[zone_idx] <- stats$sd_SAND
}

# Create spatial visualization
par(mfrow = c(1, 2))

# Prediction map
x_unique <- sort(unique(eval_grid$x))
y_unique <- sort(unique(eval_grid$y))
pred_matrix <- matrix(eval_grid$pred_sand, 
                      nrow = length(x_unique), 
                      ncol = length(y_unique))

image(x_unique, y_unique, pred_matrix,
      main = "Predicted SAND by Zone",
      xlab = "Easting", ylab = "Northing",
      col = terrain.colors(20))
points(all_data$x, all_data$y, pch = 3, cex = 0.8)

# Uncertainty map
sd_matrix <- matrix(eval_grid$pred_sd, 
                    nrow = length(x_unique), 
                    ncol = length(y_unique))

image(x_unique, y_unique, sd_matrix,
      main = "Ensemble Uncertainty (SD)",
      xlab = "Easting", ylab = "Northing",
      col = heat.colors(20))
points(all_data$x, all_data$y, pch = 3, cex = 0.8)
```

