---
title: "Per-Zone Risk Assessment & Decision Support"
author: "Andrew G. Brown"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_document:
    number_sections: true
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 10,
  fig.height = 7,
  dpi = 100,
  eval = FALSE
)
```

# Introduction: Decision-Making Under Asymmetric Costs

**Objective**: Translate zone-stratified uncertainty into actionable decisions using expected loss frameworks.

**This example covers:**

1. **Expected loss computation** — Cost-benefit analysis per zone
2. **Asymmetric cost functions** — Different penalties for over/under-application
3. **Carbon compliance auditing** — Carbon stock verification with uncertainty quantification
4. **Decision prioritization** — Which zones need resources first?

**Prerequisites:**
- Familiarity with stratified hierarchical modeling (Example 03)
- Understand per-zone ensemble aggregation (Example 04)
- Basic understanding of decision theory

**Why Per-Zone Risk Assessment Matters:**

Standard risk metrics (e.g., "P(SAND < 50%) = 30%") ignore context:
- A farmer in upland may accept high clay (landscape naturally clay-rich)
- A farmer in lowland must avoid high clay (drainage constraints)
- A carbon auditor must quantify uncertainty for compliance reporting
- Land managers need zone-specific cost-benefit tradeoffs

Per-zone decision support reveals:
- Which zones pose highest financial/environmental risk?
- Where are resources most cost-effective?
- Which zones have sufficient carbon verification for credit issuance?

---

# Setup & Data Preparation

```{r libraries}
library(geocoda)
library(terra)
library(sf)

set.seed(42)
```

## Create Stratified Soil Data with Zone-Specific Characteristics

```{r generate_data}
# Three zones with different compositions and sample sizes
# Upland: sandy, well-drained
upland <- data.frame(
  zone = "Upland",
  x = runif(15, 0, 30),
  y = runif(15, 0, 30),
  SAND = pmax(60, pmin(85, rnorm(15, mean = 72, sd = 8))),
  SILT = NA,
  CLAY = NA
)

# Midslope: loamy, transition
midslope <- data.frame(
  zone = "Midslope",
  x = runif(25, 35, 65),
  y = runif(25, 35, 65),
  SAND = pmax(35, pmin(60, rnorm(25, mean = 48, sd = 10))),
  SILT = NA,
  CLAY = NA
)

# Lowland: clay-rich, poorly drained
lowland <- data.frame(
  zone = "Lowland",
  x = runif(10, 70, 100),
  y = runif(10, 70, 100),
  SAND = pmax(20, pmin(40, rnorm(10, mean = 28, sd = 6))),
  SILT = NA,
  CLAY = NA
)

all_data <- rbind(upland, midslope, lowland)

# Complete composition to close at 100%
all_data$SILT <- 30 - all_data$SAND * 0.2
all_data$CLAY <- 100 - all_data$SAND - all_data$SILT
all_data$CLAY <- pmax(0, pmin(100 - all_data$SAND - all_data$SILT, 100 - all_data$SAND - all_data$SILT))

# Verify closure
cat("Data coverage by zone:\n")
print(table(all_data$zone))
cat("\nComposition summary (mean %):\n")
print(aggregate(all_data[, c("SAND", "SILT", "CLAY")], 
                by = list(Zone = all_data$zone), 
                FUN = mean))
```

---

# Step 1: Fit Stratified Hierarchical Model

Quick HZM fit on stratified zones:

```{r fit_model}
# Define hierarchy
hierarchy <- gc_define_hierarchy(
  data = all_data[, c("SAND", "SILT", "CLAY")],
  zone_col = "zone",
  zone_vector = all_data$zone,
  levels = c("zone", "global")
)

# Set priors
priors <- gc_set_hierarchy_priors(
  data = all_data[, c("SAND", "SILT", "CLAY")],
  hierarchy = hierarchy,
  mean_strength = 1.5,
  adjust_strength_by_n = TRUE
)

# Fit HZM
hzm_fit <- gc_fit_hierarchical_model(
  data = all_data[, c("SAND", "SILT", "CLAY")],
  hierarchy = hierarchy,
  priors = priors,
  backend = "analytical",
  verbose = FALSE
)

cat("Model fitted successfully. Generating ensemble...\n")
```

---

# Step 2: Generate Ensemble & Per-Zone Statistics

```{r generate_ensemble}
# Generate ensemble realizations
n_realizations <- 100

ensemble <- gc_sim_hierarchical(
  object = hzm_fit,
  n = n_realizations,
  zone_vector = all_data$zone,
  coords = all_data[, c("x", "y")],
  method = "turn_based"
)

cat("Ensemble generated:\n")
cat("  Realizations:", n_realizations, "\n")
cat("  Locations:", nrow(ensemble$data), "\n")

# Compute per-zone ensemble statistics
per_zone_stats <- gc_ensemble_per_zone(
  ensemble = ensemble,
  zone_col = "zone_id",
  verbose = FALSE
)

cat("\nPer-zone summary (SAND):\n")
print(per_zone_stats$zone_stats[, c("zone", "n", "SAND_mean", "SAND_sd", "SAND_lower", "SAND_upper")])
```

---

# Step 3: Define Risk Thresholds & Cost Functions

Classic farmer decision scenario: management constraints and asymmetric costs.

```{r define_costs}
# Thresholds (critical values)
# - SAND < 40%: Insufficient drainage (problem)
# - CLAY > 35%: Compaction risk (problem)

thresholds <- c(
  SAND = 40,  # Minimum sand for drainage
  CLAY = 35   # Maximum clay before compaction
)

# Asymmetric cost matrix
# Units: arbitrary economic loss units (could be $ or crop yield loss)
cost_matrix <- data.frame(
  component = c("SAND", "CLAY"),
  cost_under = c(150, 0),      # SAND < 40% costs 150 (drainage failure = high cost)
                                # CLAY < 35% has zero cost (lower is better)
  cost_over = c(0, 120)         # SAND > 40% has zero cost (higher is fine)
                                 # CLAY > 35% costs 120 (compaction risk)
)

cat("Risk Threshold & Cost Matrix:\n")
cat("Thresholds:\n")
print(thresholds)
cat("\nCost Matrix (economic loss units):\n")
print(cost_matrix)
```

---

# Step 4: Compute Per-Zone Expected Loss

```{r per_zone_risk}
# Compute expected loss by zone
risk_results <- gc_risk_assessment_per_zone(
  ensemble = ensemble,
  zone_col = "zone_id",
  thresholds = thresholds,
  cost_matrix = cost_matrix,
  verbose = TRUE
)

cat("\n\nExpected Loss Summary:\n")
print(risk_results)
```

---

# Step 5: Interpret Results & Prioritize Management Actions

```{r interpret_risk}
# Summarize by zone
zone_risk_summary <- aggregate(
  expected_loss ~ zone,
  data = risk_results,
  FUN = function(x) c(mean = mean(x), max = max(x))
)

cat("Zone Risk Ranking (mean expected loss):\n")
zone_summary <- data.frame(
  zone = unique(risk_results$zone),
  mean_expected_loss = aggregate(expected_loss ~ zone, data = risk_results, FUN = mean)$expected_loss,
  max_expected_loss = aggregate(expected_loss ~ zone, data = risk_results, FUN = max)$expected_loss
)

zone_summary <- zone_summary[order(zone_summary$mean_expected_loss, decreasing = TRUE), ]
print(zone_summary)

cat("\n\nInterpretation:\n")
cat("High expected loss zones:\n")
high_risk_zones <- zone_summary[zone_summary$mean_expected_loss > median(zone_summary$mean_expected_loss), ]
for (i in seq_len(nrow(high_risk_zones))) {
  zone_name <- high_risk_zones[i, 1]
  loss <- high_risk_zones[i, 2]
  cat(sprintf("  %s: EL = %.1f (prioritize mitigation)\n", zone_name, loss))
}
```

---

# Step 6: Carbon Compliance Audit

Compute per-zone carbon stock with compliance verification:

```{r carbon_audit}
# Audit carbon stock by zone
carbon_audit <- gc_carbon_audit_per_zone(
  ensemble = ensemble,
  zone_col = "zone_id",
  soil_depth = 30,        # Audit top 30cm
  bulk_density = NULL,    # Use default by texture
  carbon_fraction = "soilgrids",  # Empirical SoilGrids C model
  credit_price = 25,      # USD per Mg C/ha
  verbose = TRUE
)

cat("\n\nCarbon Stock Results:\n")
print(carbon_audit)
```

---

# Step 7: Decision Matrix & Resource Allocation

Combine risk and carbon audit for comprehensive zone prioritization:

```{r decision_matrix}
# Create decision matrix combining risk and carbon metrics
decision_table <- data.frame(
  Zone = zone_summary$zone,
  Avg_Expected_Loss = zone_summary$mean_expected_loss,
  Risk_Ranking = rank(-zone_summary$mean_expected_loss),
  C_Stock_Verified = carbon_audit$compliance_status[
    match(zone_summary$zone, carbon_audit$zone)
  ],
  C_Credit_USD = carbon_audit$credit_value_usd[
    match(zone_summary$zone, carbon_audit$zone)
  ]
)

decision_table$Overall_Priority <- ifelse(
  decision_table$Risk_Ranking <= 2 & decision_table$C_Stock_Verified != "Verified",
  "High (Risk + Carbon Uncertainty)",
  ifelse(
    decision_table$Risk_Ranking <= 2,
    "High (Risk Only)",
    "Medium/Low"
  )
)

cat("Decision & Resource Allocation Matrix:\n")
print(decision_table)

cat("\n\nRecommendations:\n")
for (i in seq_len(nrow(decision_table))) {
  zone_name <- decision_table$Zone[i]
  priority <- decision_table$Overall_Priority[i]
  carbon_status <- decision_table$C_Stock_Verified[i]
  
  cat(sprintf("\"%s\" zone: ", zone_name))
  cat(sprintf("Priority = %s\n", priority))
  
  if (carbon_status != "Verified") {
    cat(sprintf("  → Recommend additional sampling for carbon verification\n"))
  }
  cat(sprintf("  → Review management practices to reduce expected loss\n\n"))
}
```

---

# Step 8: Visualization

Box plots showing risk distribution and carbon uncertainty:

```{r visualization}
par(mfrow = c(2, 2), mar = c(4, 4, 2, 1))

# 1. Expected Loss by Zone & Component
sand_risk <- risk_results[risk_results$component == "SAND", ]
boxplot(
  list(
    Upland = sand_risk$expected_loss[sand_risk$zone == "Upland"],
    Midslope = sand_risk$expected_loss[sand_risk$zone == "Midslope"],
    Lowland = sand_risk$expected_loss[sand_risk$zone == "Lowland"]
  ),
  main = "Expected Loss (SAND drainage threshold)",
  ylab = "Expected Loss (units)",
  col = c("tan", "lightgreen", "lightblue")
)

# 2. Carbon Stock by Zone
carbon_audit_sorted <- carbon_audit[
  match(c("Upland", "Midslope", "Lowland"), carbon_audit$zone),
]

barplot(
  carbon_audit_sorted$mean_c_stock_mg_ha,
  names.arg = carbon_audit_sorted$zone,
  main = "Mean Carbon Stock by Zone",
  ylab = "C Stock (Mg/ha)",
  ylim = c(0, max(carbon_audit_sorted$ci_upper) * 1.2),
  col = c("tan", "lightgreen", "lightblue")
)

# Add error bars (manual)
mids <- 1:3
segments(mids, carbon_audit_sorted$ci_lower, mids, carbon_audit_sorted$ci_upper)
segments(mids - 0.1, carbon_audit_sorted$ci_lower, mids + 0.1, carbon_audit_sorted$ci_lower)
segments(mids - 0.1, carbon_audit_sorted$ci_upper, mids + 0.1, carbon_audit_sorted$ci_upper)

# 3. Carbon Compliance Status
compliance_counts <- table(carbon_audit$zone, carbon_audit$compliance_status)
barplot(
  t(compliance_counts)[, c("Upland", "Midslope", "Lowland")],
  main = "Carbon Compliance Status by Zone",
  ylab = "Count",
  legend.text = TRUE,
  col = c("darkred", "orange", "darkgreen")
)

# 4. Risk vs Carbon Trade-off
plot(
  zone_summary$mean_expected_loss,
  carbon_audit_sorted$mean_c_stock_mg_ha,
  main = "Risk vs Carbon Stock (trade-off analysis)",
  xlab = "Expected Loss",
  ylab = "C Stock (Mg/ha)",
  pch = 16,
  cex = 2,
  col = c("tan", "lightgreen", "lightblue")[seq_len(nrow(zone_summary))]
)

text(
  zone_summary$mean_expected_loss,
  carbon_audit_sorted$mean_c_stock_mg_ha,
  labels = c("Upland", "Midslope", "Lowland"),
  pos = 3,
  cex = 0.9
)
```

---

# Summary & Best Practices

## When to Use Per-Zone Risk Assessment

Use this approach when:

1. **Asymmetric costs**: Over- and under-application have different financial impacts
2. **Zone heterogeneity**: Zones have naturally different "acceptable" ranges
3. **Resource-constrained**: Limited budget requires prioritization
4. **Compliance-driven**: Carbon audits, environmental regulations require per-zone verification

## Key Insights

- **Upland zones** (high sand): Drainage not a concern; risk from CLAY > 35% is minimal (sandy soils naturally resistant)
- **Midslope zones** (loamy): Mixed concerns; requires balanced assessment
- **Lowland zones** (clay-rich): High risk from compaction; per-zone decision support critical

## Decision Framework

| Metric | Decision |
|--------|----------|
| High expected loss + wide C CI | Highest priority: manage risk + sample carbon |
| High expected loss + verified C | High priority: manage risk, confidence in carbon |
| Low expected loss + uncertain C | Medium priority: verify carbon for compliance |
| Low expected loss + verified C | Lowest priority: maintain current practices |

---

# Advanced: Sensitivity Analysis (Optional)

```{r sensitivity, eval=FALSE}
# How does changing thresholds affect zone risk rankings?

# Test alternative thresholds
thresholds_optimistic <- c(SAND = 35, CLAY = 40)
thresholds_conservative <- c(SAND = 50, CLAY = 30)

risk_optimistic <- gc_risk_assessment_per_zone(
  ensemble, zone_col = "zone_id",
  thresholds = thresholds_optimistic,
  cost_matrix = cost_matrix, verbose = FALSE
)

risk_conservative <- gc_risk_assessment_per_zone(
  ensemble, zone_col = "zone_id",
  thresholds = thresholds_conservative,
  cost_matrix = cost_matrix, verbose = FALSE
)

# Compare rankings
cat("Sensitivity to threshold choice:\n\n")
cat("Central thresholds (SAND=40, CLAY=35):\n")
print(aggregate(expected_loss ~ zone, data = risk_results, FUN = mean))

cat("\nOptimistic (SAND=35, CLAY=40):\n")
print(aggregate(expected_loss ~ zone, data = risk_optimistic, FUN = mean))

cat("\nConservative (SAND=50, CLAY=30):\n")
print(aggregate(expected_loss ~ zone, data = risk_conservative, FUN = mean))
```

