---
title: "Zone-Stratified Model Diagnostics & Uncertainty Quantification"
author: "geocoda"
date: "`r Sys.Date()`"
output: rmarkdown::html_document
vignette: >
  %\VignetteIndexEntry{Zone-Stratified Model Diagnostics & Uncertainty Quantification}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 10,
  fig.height = 6,
  dpi = 100
)
```

# Zone-Stratified Model Diagnostics & Uncertainty Quantification

This example demonstrates comprehensive model diagnostics stratified by zone.
Identifies under-validated zones needing additional data, quantifies compositional
uncertainty hotspots, and assesses parameter stability under bootstrap resampling.

## Setup: Create Three-Zone Domain

```{r setup-data}
library(geocoda)
library(ggplot2)
library(sf)

# Create synthetic three-zone domain
set.seed(42)
n_samples <- 150

# Zone assignments: upland (1), midslope (2), lowland (3)
zone_vector <- rep(c("upland", "midslope", "lowland"), each = n_samples / 3)

# Compositional data with zone-specific mean compositions
comps <- data.frame(
  # Upland: sandy (high SAND variability, low CLAY)
  SAND = c(
    rnorm(n_samples / 3, mean = 75, sd = 8),
    rnorm(n_samples / 3, mean = 60, sd = 10),
    rnorm(n_samples / 3, mean = 45, sd = 12)
  ),
  # Midslope: loamy (balanced)
  SILT = c(
    rnorm(n_samples / 3, mean = 15, sd = 5),
    rnorm(n_samples / 3, mean = 25, sd = 8),
    rnorm(n_samples / 3, mean = 30, sd = 10)
  ),
  # Lowland: clayey (high CLAY, low SAND)
  CLAY = c(
    rnorm(n_samples / 3, mean = 10, sd = 4),
    rnorm(n_samples / 3, mean = 15, sd = 6),
    rnorm(n_samples / 3, mean = 25, sd = 8)
  ),
  zone = zone_vector
)

# Closure constraint: normalize to 100
comps[, 1:3] <- prop.table(as.matrix(comps[, 1:3]), margin = 1) * 100

head(comps, 10)
```

## 1. Fit Hierarchical Model

```{r fit-model}
# Define hierarchy
hz_def <- gc_define_hierarchy(
  data = comps,
  comp_cols = c("SAND", "SILT", "CLAY"),
  group_col = "zone",
  var_order = "variance"
)

# Fit HZM
hz_fit <- gc_fit_hierarchical_model(
  data = comps,
  object = hz_def,
  verbose = FALSE
)

cat("Hierarchical model fitted.\n")
cat("Structure:\n")
print(hz_fit$hierarchy)
```

## 2. Per-Zone Cross-Validation

Assess model predictiveness stratified by zone.

```{r cv-per-zone}
cv_results <- gc_cross_validate_per_zone(
  data = comps,
  zone_vector = zone_vector,
  comp_cols = c("SAND", "SILT", "CLAY"),
  k = 5,
  verbose = TRUE
)

cat("\n=== Cross-Validation Summary RMSE by Zone ===\n")
print(cv_results[, c("zone", "n_samples", "mean_rmse", "sd_rmse", "mean_r_squared")])

# Visualize CV results
cv_plot <- ggplot(cv_results, aes(x = reorder(zone, -mean_rmse), y = mean_rmse, 
                                   fill = zone)) +
  geom_col() +
  geom_errorbar(aes(ymin = mean_rmse - sd_rmse, ymax = mean_rmse + sd_rmse), 
                width = 0.2) +
  labs(
    title = "Cross-Validation RMSE by Zone",
    x = "Zone",
    y = "Mean RMSE (± 1 SD)",
    subtitle = "Lower RMSE indicates better model fit"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

print(cv_plot)
```

**Interpretation**:
- **Upland** (low RMSE): Well-fitted, sufficient data
- **Midslope** (moderate RMSE): Adequate fit
- **Lowland** (high RMSE): Underfitted; may need more samples

## 3. Generate Ensemble

```{r generate-ensemble}
# Simulate hierarchical ensemble
ensemble <- gc_sim_hierarchical(
  object = hz_fit,
  n = 500,
  zone_vector = comps$zone,
  coords = data.frame(x = seq_len(nrow(comps)), y = rep(1, nrow(comps)))
)

cat("Ensemble generated:\n")
cat("  Realizations:", length(ensemble$data[[1]]), "\n")
cat("  Zones:", length(unique(ensemble$data$zone_id)), "\n")
```

## 4. Per-Zone Compositional Entropy

Identify high-uncertainty zones (compositional variability hotspots).

```{r entropy-per-zone}
entropy_results <- gc_compute_entropy_per_zone(
  ensemble = ensemble,
  zone_col = "zone_id",
  method = "shannon",
  normalize = TRUE,
  verbose = TRUE
)

cat("\n=== Entropy Summary ===\n")
print(entropy_results[, c("zone", "n_ensemble", "shannon_entropy", 
                          "dominant_component", "uncertainty_class")])

# Visualize entropy
entropy_plot <- ggplot(entropy_results, aes(x = reorder(zone, -shannon_entropy), 
                                            y = shannon_entropy, 
                                            fill = uncertainty_class)) +
  geom_col() +
  scale_fill_manual(values = c("Low" = "green", "Medium" = "yellow", "High" = "red")) +
  labs(
    title = "Compositional Entropy by Zone",
    x = "Zone",
    y = "Shannon Entropy (normalized)",
    subtitle = "High entropy = greater compositional uncertainty",
    fill = "Uncertainty"
  ) +
  theme_minimal()

print(entropy_plot)
```

**Interpretation**:
- **Low Entropy** (e.g., upland): Homogeneous composition (low uncertainty)
- **High Entropy** (e.g., lowland): Diverse composition (high uncertainty, may need targeted sampling)

## 5. Bootstrap Parameter Uncertainty

Assess robustness of zone-specific parameter estimates.

```{r bootstrap-per-zone}
bootstrap_results <- gc_bootstrap_uncertainty_per_zone(
  data = comps,
  zone_vector = comps$zone,
  n_bootstrap = 1000,
  conf_level = 0.95,
  verbose = TRUE
)

cat("\n=== Parameter CI Summary ===\n")
print(bootstrap_results$zone_estimates[, c("zone", "n_obs", "param_1_est", 
                                           "param_1_lower", "param_1_upper")])
```

## 6. Combined Diagnostic Dashboard

Integrate all diagnostics into a single assessment.

```{r diagnostic-dashboard}
# Create summary data frame combining all metrics
diagnostic_summary <- data.frame(
  zone = cv_results$zone,
  model_fit_rmse = cv_results$mean_rmse,
  cv_stability = cv_results$sd_rmse,
  composition_entropy = entropy_results$shannon_entropy,
  uncertainty_class = entropy_results$uncertainty_class,
  param_uncertainty = bootstrap_results$zone_estimates$param_1_upper - 
                      bootstrap_results$zone_estimates$param_1_lower
)

cat("=== Comprehensive Zone Diagnostic Summary ===\n\n")
print(diagnostic_summary)

# Diagnostic recommendation matrix
cat("\n=== Diagnostic Recommendations ===\n\n")
for (i in seq_len(nrow(diagnostic_summary))) {
  zone_name <- diagnostic_summary$zone[i]
  rmse <- diagnostic_summary$model_fit_rmse[i]
  entropy <- diagnostic_summary$composition_entropy[i]
  param_unc <- diagnostic_summary$param_uncertainty[i]
  
  cat(sprintf("Zone: %s\n", zone_name))
  
  if (rmse > mean(diagnostic_summary$model_fit_rmse) * 1.3) {
    cat("  ⚠ Model fit is poor relative to other zones\n")
    cat("    Recommendation: Increase sampling, check for outliers\n")
  } else {
    cat("  ✓ Model fit is adequate\n")
  }
  
  if (entropy > quantile(diagnostic_summary$composition_entropy, 0.75)) {
    cat("  ⚠ High compositional entropy detected\n")
    cat("    Recommendation: Target uncertain compositions for more data\n")
  } else {
    cat("  ✓ Composition is relatively homogeneous\n")
  }
  
  if (param_unc > quantile(diagnostic_summary$param_uncertainty, 0.75)) {
    cat("  ⚠ Parameter estimates have wide confidence intervals\n")
    cat("    Recommendation: More data needed for stable estimates\n")
  } else {
    cat("  ✓ Parameter estimates are stable\n")
  }
  
  cat("\n")
}
```

## 7. Advanced: Multi-Diagnostic Composite Assessment

```{r composite-assessment}
# Normalize metrics to [0, 1] for composite scoring
diagnostic_summary$rmse_norm <- 
  (diagnostic_summary$model_fit_rmse - min(diagnostic_summary$model_fit_rmse)) /
  (max(diagnostic_summary$model_fit_rmse) - min(diagnostic_summary$model_fit_rmse))

diagnostic_summary$entropy_norm <-
  (diagnostic_summary$composition_entropy - min(diagnostic_summary$composition_entropy)) /
  (max(diagnostic_summary$composition_entropy) - min(diagnostic_summary$composition_entropy))

diagnostic_summary$param_unc_norm <-
  (diagnostic_summary$param_uncertainty - min(diagnostic_summary$param_uncertainty)) /
  (max(diagnostic_summary$param_uncertainty) - min(diagnostic_summary$param_uncertainty))

# Composite "confidence" score (inverse of uncertainty)
diagnostic_summary$confidence_score <- 1 - (
  0.4 * diagnostic_summary$rmse_norm +
  0.3 * diagnostic_summary$entropy_norm +
  0.3 * diagnostic_summary$param_unc_norm
)

cat("=== Zone Confidence Scores (0-1, higher is better) ===\n")
print(diagnostic_summary[, c("zone", "confidence_score")])

# Visualize composite score
confidence_plot <- ggplot(diagnostic_summary, aes(x = reorder(zone, -confidence_score), 
                                                   y = confidence_score, 
                                                   fill = confidence_score)) +
  geom_col() +
  scale_fill_gradient(low = "red", high = "green", limits = c(0, 1)) +
  labs(
    title = "Zone Model Confidence Score",
    x = "Zone",
    y = "Confidence (0-1)",
    subtitle = "Composite of model fit, composition stability, parameter precision"
  ) +
  theme_minimal() +
  theme(legend.position = "right")

print(confidence_plot)
```

## 8. Best Practices

### Cross-Validation Interpretation

- **RMSE < 5%**: Excellent zone-specific predictions
- **RMSE 5-10%**: Good predictions, acceptable for most applications
- **RMSE > 10%**: Poor predictions, zone needs more data

### Entropy Hotspots

High-entropy zones indicate:
1. Natural compositional variability (no action needed)
2. Insufficient sampling (add targeted samples)
3. Multiple sub-domains (consider stratifying further)

### Parameter Stability

- If parameter CI overlaps between zones, they may be mergeable
- If parameter CI non-overlapping, zones are distinct (keep separate)

## Summary

The per-zone diagnostic workflow enables:

1. **Data Gap Identification**: Detect under-sampled zones via CV and entropy
2. **Uncertainty Quantification**: Assess parameter stability via bootstrap
3. **Sampling Design**: Recommend where to collect additional data
4. **Model Validation**: Verify zone appropriateness and distinctness
5. **Confidence Assessment**: Composite score combining all metrics

Use these diagnostics to iteratively refine your zone definitions and sampling strategy.
