---
title: "Stratified 3D Geostatistics: Lateral Domains × Depth Strata"
author: "geocoda"
date: "`r Sys.Date()`"
output: rmarkdown::html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 11,
  fig.height = 7,
  dpi = 100
)
```

# Stratified 3D Geostatistics

This example demonstrates 3D geostatistical simulation respecting both lateral
domain stratification and vertical depth structure. Combines domain-specific
means with depth-dependent variability.

## 1. Create Synthetic 3D Dataset

```{r create-3d-data}
library(geocoda)
library(ggplot2)
library(reshape2)

set.seed(42)

# Create three lateral zones in 100×100 km domain
# Zone 1: Upland (x < 33)
# Zone 2: Midslope (33 ≤ x < 67)
# Zone 3: Lowland (x ≥ 67)

n_samples <- 300
zone_assignments <- cut(sample.int(100, n_samples, replace = TRUE),
                        breaks = c(0, 33, 67, 100),
                        labels = c("upland", "midslope", "lowland"))

# 3D coordinates
data_3d <- data.frame(
  x = sample.int(100, n_samples, replace = TRUE),
  y = sample.int(100, n_samples, replace = TRUE),
  z = sample.int(100, n_samples, replace = TRUE), # depth 0-100 cm
  zone_id = zone_assignments
)

# Zone-specific compositions (varying with depth)
set.seed(42)
for (i in seq_along(data_3d$zone_id)) {
  zone <- as.character(data_3d$zone_id[i])
  depth <- data_3d$z[i]
  depth_factor <- 1 - (0.3 * depth / 100) # decrease variability with depth

  if (zone == "upland") {
    # Sandy uplands
    sand <- rnorm(1, mean = 75, sd = 8 * depth_factor)
    silt <- rnorm(1, mean = 15, sd = 5 * depth_factor)
    clay <- rnorm(1, mean = 10, sd = 3 * depth_factor)
  } else if (zone == "midslope") {
    # Loamy midslopes
    sand <- rnorm(1, mean = 55, sd = 10 * depth_factor)
    silt <- rnorm(1, mean = 25, sd = 8 * depth_factor)
    clay <- rnorm(1, mean = 20, sd = 5 * depth_factor)
  } else { # lowland
    # Clayey lowlands
    sand <- rnorm(1, mean = 35, sd = 12 * depth_factor)
    silt <- rnorm(1, mean = 30, sd = 10 * depth_factor)
    clay <- rnorm(1, mean = 35, sd = 8 * depth_factor)
  }

  # Closure
  total <- sand + silt + clay
  data_3d$SAND[i] <- 100 * sand / total
  data_3d$SILT[i] <- 100 * silt / total
  data_3d$CLAY[i] <- 100 * clay / total
}

cat("3D Dataset Summary\n")
cat("==================\n")
cat(sprintf("Total samples: %d\n", nrow(data_3d)))
cat(sprintf("Zones: %s\n", paste(unique(data_3d$zone_id), collapse = ", ")))
cat(sprintf("Depth range: %d - %d cm\n", min(data_3d$z), max(data_3d$z)))
cat("\nZone-specific composition profiles:\n")
for (zone in unique(data_3d$zone_id)) {
  zone_data <- data_3d[data_3d$zone_id == zone, ]
  cat(sprintf("  %s: SAND=%.1f±%.1f, SILT=%.1f±%.1f, CLAY=%.1f±%.1f\n",
              zone,
              mean(zone_data$SAND), sd(zone_data$SAND),
              mean(zone_data$SILT), sd(zone_data$SILT),
              mean(zone_data$CLAY), sd(zone_data$CLAY)))
}
```

## 2. Define Nested Hierarchy (Lateral × Depth)

```{r define-nested-hierarchy}
# Define hierarchy: 3 lateral zones × 3 depth strata = 9 hierarchical units
h_nested <- gc_define_nested_hierarchy(
  domain_zones = c("upland", "midslope", "lowland"),
  depth_strata = c("shallow", "mid", "deep"),
  depth_breaks = c(0, 30, 60, 100)
)

cat("Nested Hierarchy Structure\n")
cat("==========================\n")
cat(sprintf("Lateral domains: %s\n", paste(h_nested$domain_zones, collapse = ", ")))
cat(sprintf("Depth strata: %s\n", paste(h_nested$depth_strata, collapse = ", ")))
cat(sprintf("Total hierarchical units: %d\n\n", h_nested$n_units))
print(h_nested$nested_units)
```

## 3. Fit Zone-Specific 3D Variograms

```{r fit-3d-variograms}
vgm_3d <- gc_fit_vgm_3d_per_zone(
  data = data_3d,
  zone_col = "zone_id",
  value_col = "SAND",
  model = "Sph"
)

cat("3D Variogram Summary by Zone\n")
cat("============================\n\n")
print("Zone Ranges:")
print(vgm_3d$zone_ranges)

cat("\n\nData Summary:\n")
print(vgm_3d$data_summary)

cat("\n\nInterpretation:\n")
cat("- Lateral range: horizontal correlation distance\n")
cat("- Vertical range: depth-dependent correlation distance\n")
cat("- Anisotropy ratio: lateral_range / vertical_range\n")
cat("  (ratio > 1: lateral more correlated than vertical)\n")
```

## 4. Fit Hierarchical Model

```{r fit-hierarchical-model-3d}
# Define hierarchy by zone only (HZM doesn't directly handle depth yet)
hz_def <- gc_define_hierarchy(
  data = data_3d,
  comp_cols = c("SAND", "SILT", "CLAY"),
  group_col = "zone_id",
  var_order = "variance"
)

# Fit HZM
hz_fit_3d <- gc_fit_hierarchical_model(
  data = data_3d,
  object = hz_def,
  verbose = FALSE
)

cat("Hierarchical Model Fitted\n")
cat("=========================\n")
print(hz_fit_3d$hierarchy)
```

## 5. Get 3D Prediction Locations (Prediction Grid)

```{r create-prediction-grid}
# Create 2D grid for prediction
grid_2d <- expand.grid(
  x = seq(10, 90, by = 10),
  y = seq(10, 90, by = 10)
)

# Add depth profile: predict at multiple depths
grid_3d <- data.frame()
for (depth in c(10, 30, 50, 75, 100)) {
  grid_depth <- grid_2d
  grid_depth$z <- depth
  grid_3d <- rbind(grid_3d, grid_depth)
}

# Assign zones to grid points
grid_3d$zone_id <- cut(grid_3d$x,
                       breaks = c(0, 33, 67, 100),
                       labels = c("upland", "midslope", "lowland"))

cat("Prediction Grid\n")
cat("===============\n")
cat(sprintf("Total grid points: %d\n", nrow(grid_3d)))
cat(sprintf("Grid extent: x=[%d,%d], y=[%d,%d], z=[%d,%d]\n",
            min(grid_3d$x), max(grid_3d$x),
            min(grid_3d$y), max(grid_3d$y),
            min(grid_3d$z), max(grid_3d$z)))
```

## 6. Simulate 3D Hierarchical Ensemble

```{r simulate-3d-ensemble}
ensemble_3d <- gc_sim_hierarchical_3d_per_zone(
  object = hz_fit_3d,
  n = 100,
  nested_hierarchy = h_nested,
  coords = grid_3d,
  zone_vector = grid_3d$zone_id
)

cat("3D Ensemble Generated\n")
cat("====================\n")
cat(sprintf("Realizations: %d\n", ensemble_3d$n_realizations))
cat(sprintf("Locations per realization: %d\n", ensemble_3d$n_locations))
cat(sprintf("Total ensemble points: %d\n\n", nrow(ensemble_3d$data)))

# Check composition closure
head(ensemble_3d$data)
```

## 7. Extract and Visualize Depth Profiles

```{r depth-profiles}
# Extract mean composition at different depths by zone
profile_data <- data.frame()

for (zone in c("upland", "midslope", "lowland")) {
  for (depth in c(10, 30, 50, 75, 100)) {
    subset <- ensemble_3d$data[
      ensemble_3d$data$zone_id == zone & ensemble_3d$data$z == depth,
    ]

    if (nrow(subset) > 0) {
      profile_data <- rbind(profile_data, data.frame(
        zone = zone,
        depth = depth,
        SAND = mean(subset$SAND),
        SILT = mean(subset$SILT),
        CLAY = mean(subset$CLAY)
      ))
    }
  }
}

cat("Zone Composition Profiles by Depth\n")
cat("==================================\n")
print(profile_data)

# Visualize
profile_long <- melt(profile_data, id.vars = c("zone", "depth"))

p_profile <- ggplot(profile_long, aes(x = depth, y = value, color = variable, 
                                      linetype = zone)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  facet_wrap(~variable) +
  scale_x_reverse() +
  coord_flip() +
  labs(
    title = "Compositional Depth Profiles by Zone",
    x = "Depth (cm)",
    y = "Proportion (%)",
    color = "Component",
    linetype = "Zone"
  ) +
  theme_minimal()

print(p_profile)
```

## 8. Spatial Variability by Depth

```{r spatial-variability-by-depth}
# Compare lateral variability across depths
variability_by_depth <- data.frame()

for (depth in unique(ensemble_3d$data$z)) {
  subset <- ensemble_3d$data[ensemble_3d$data$z == depth, ]

  for (zone in unique(subset$zone_id)) {
    zone_subset <- subset[subset$zone_id == zone, ]

    # Compute coefficient of variation
    cv_sand <- sd(zone_subset$SAND, na.rm = TRUE) / mean(zone_subset$SAND, na.rm = TRUE)

    variability_by_depth <- rbind(variability_by_depth, data.frame(
      depth = depth,
      zone = zone,
      cv_sand = cv_sand,
      mean_sand = mean(zone_subset$SAND, na.rm = TRUE)
    ))
  }
}

p_variability <- ggplot(variability_by_depth, 
                        aes(x = depth, y = cv_sand, color = zone, shape = zone)) +
  geom_point(size = 3) +
  geom_line(alpha = 0.5) +
  scale_x_reverse() +
  coord_flip() +
  labs(
    title = "SAND Content Variability by Depth",
    x = "Depth (cm)",
    y = "Coefficient of Variation",
    color = "Zone",
    shape = "Zone"
  ) +
  theme_minimal()

print(p_variability)

cat("\nNote: CV tends to decrease with depth (compaction effect)\n")
```

## 9. Best Practices for 3D Stratified Modeling

### Zone Definition in 3D

- **Lateral stratification**: Based on topography, geology, land use
- **Vertical stratification**: Based on soil horizons, pedogenic processes
- **Interaction**: Different zones may have distinct depth profiles

### Variogram Fitting

1. Ensure sufficient samples per zone (≥20)
2. Compute 3D distances (not just 2D horizontal)
3. Account for anisotropy (lateral ≠ vertical)
4. Separate variograms per zone prevent over-smoothing

### Hierarchical Modeling

1. Define zones/strata meaningfully
2. Fit HZM to account for multi-scale structure
3. Simulate ensemble with depth-dependent variability
4. Validate via cross-validation (Phase 3.1)

### Ensemble Aggregation

- Compute **per-zone statistics** (not global means)
- Use **depth-stratified summaries** for interpretation
- Map **uncertainty hotspots** (high entropy or wide CIs)

## Summary

3D stratified geostatistical modeling enables:

1. **Domain-aware simulation**: Each zone gets zone-specific params
2. **Depth integration**: Respects pedogenic processes (compaction, leaching)
3. **Spatial correlation**: 3D variograms capture lateral + vertical structure
4. **Ensemble-based inference**: Quantify uncertainty at any depth/location

Use this workflow as a foundation for multi-scale risk assessment and resource mapping.
