---
title: "Complete 3D Stratified Workflow: Mapping and Decision Support"
author: "geocoda"
date: "`r Sys.Date()`"
output: rmarkdown::html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 12,
  fig.height = 8,
  dpi = 100
)
```

# Complete 3D Stratified Geostatistical Workflow

This example demonstrates an end-to-end 3D workflow: from domain stratification
through 3D ensemble simulation to uncertainty mapping and decision support.
Applies 3D simulation for resource assessment under complex domain structure.

## 1. Problem Setup: Soil Carbon Inventory in Stratified Terrain

```{r problem-setup}
library(geocoda)
library(ggplot2)
library(dplyr)

# Goal: Estimate soil carbon stocks in 3D across layered topography
# with uncertain domain boundaries

# Create realistic 3-zone domain with topographic influence
set.seed(123)

n_samples_per_zone <- 80

# Geographic zones (lateral stratification)
upland <- data.frame(
  zone = "upland",
  x = rnorm(n_samples_per_zone, mean = 2.5, sd = 1.2),
  y = rnorm(n_samples_per_zone, mean = 5, sd = 2),
  elev = rnorm(n_samples_per_zone, mean = 450, sd = 50)
)

midslope <- data.frame(
  zone = "midslope",
  x = rnorm(n_samples_per_zone, mean = 5, sd = 1.5),
  y = rnorm(n_samples_per_zone, mean = 5, sd = 2),
  elev = rnorm(n_samples_per_zone, mean = 300, sd = 40)
)

lowland <- data.frame(
  zone = "lowland",
  x = rnorm(n_samples_per_zone, mean = 7.5, sd = 1.2),
  y = rnorm(n_samples_per_zone, mean = 5, sd = 2),
  elev = rnorm(n_samples_per_zone, mean = 150, sd = 30)
)

survey_data <- rbind(upland, midslope, lowland)

# Add depth dimension (soil profiles: sampled at 3 depths)
profiles_list <- list()
for (i in seq_len(nrow(survey_data))) {
  for (depth in c("surface", "mid", "deep")) {
    depth_z <- ifelse(depth == "surface", 10,
                      ifelse(depth == "mid", 40, 75))

    if (depth == "surface") {
      oc_pct <- rnorm(1, mean = 2.5, sd = 0.5) # organic carbon %
    } else if (depth == "mid") {
      oc_pct <- rnorm(1, mean = 1.2, sd = 0.3)
    } else {
      oc_pct <- rnorm(1, mean = 0.5, sd = 0.2)
    }

    profiles_list[[length(profiles_list) + 1]] <- data.frame(
      zone = survey_data$zone[i],
      x = survey_data$x[i],
      y = survey_data$y[i],
      elev = survey_data$elev[i],
      depth_cm = depth_z,
      depth_class = depth,
      oc_percent = max(oc_pct, 0.1)
    )
  }
}

soil_profiles <- do.call(rbind, profiles_list)
rownames(soil_profiles) <- NULL

cat("Soil Profile Survey Data\n")
cat("========================\n")
cat(sprintf("Samples: %d profiles (3 depths each)\n", nrow(soil_profiles) / 3))
cat(sprintf("Zones: %s\n", paste(unique(soil_profiles$zone), collapse = ", ")))
cat(sprintf("Depth range: %d - %d cm\n", 
            min(soil_profiles$depth_cm), max(soil_profiles$depth_cm)))
cat(sprintf("\nOrganic Carbon (%%) by Zone and Depth:\n"))

oc_summary <- soil_profiles %>%
  group_by(zone, depth_class) %>%
  summarise(
    mean_oc = mean(oc_percent),
    sd_oc = sd(oc_percent),
    n = n(),
    .groups = "drop"
  )
print(oc_summary)
```

## 2. Spatial Structure Analysis

```{r spatial-structure}
# Examine spatial clustering by zone
p1 <- ggplot(soil_profiles, aes(x = x, y = elev, color = zone)) +
  geom_point(size = 2, alpha = 0.6) +
  facet_wrap(~depth_class) +
  labs(
    title = "Survey Sampling: Spatial Distribution by Depth",
    x = "X Coordinate (km)",
    y = "Elevation (m)",
    color = "Zone"
  ) +
  theme_minimal()

print(p1)

# Zone boundary uncertainty
cat("\nZone Boundary Definition\n")
cat("========================\n")
cat("Upland: x < 3.5 km\n")
cat("Midslope: 3.5 ≤ x < 6 km\n")
cat("Lowland: x ≥ 6 km\n")
cat("\nNote: Zone boundaries in this example are overlapping distributions\n")
cat("In practice, use: topography, geology, visible soil properties\n")
```

## 3. Define Nested Hierarchy and Fit HZM

```{r fit-hierarchical-model}
# Create nested hierarchy for 3D domain
h_nested <- gc_define_nested_hierarchy(
  domain_zones = c("upland", "midslope", "lowland"),
  depth_strata = c("surface", "mid", "deep"),
  depth_breaks = c(0, 25, 55, 100)
)

cat("Nested Hierarchy for 3D Modeling\n")
cat("=================================\n")
print(h_nested$nested_units)

# Create auxiliary data for HZM fitting (ILR of compositions)
# For this example, use OC as univariate; in practice would use full 3-part composition
hzm_data <- data.frame(
  OC = soil_profiles$oc_percent,
  zone = soil_profiles$zone,
  depth = soil_profiles$depth_cm,
  x = soil_profiles$x,
  y = soil_profiles$y
)

# Fit HZM by zone
hz_def <- gc_define_hierarchy(
  data = hzm_data,
  comp_cols = "OC",
  group_col = "zone",
  var_order = "variance"
)

hz_fit <- gc_fit_hierarchical_model(
  data = hzm_data,
  object = hz_def,
  verbose = FALSE
)

cat("\nHierarchical Model Summary\n")
cat("==========================\n")
print(hz_fit$hierarchy)
```

## 4. Generate 3D Prediction Domain

```{r create-prediction-domain}
# Create 3D prediction grid across domain
pred_grid <- expand.grid(
  x = seq(1, 9, by = 0.5),
  y = seq(2, 8, by = 0.5),
  depth = c(10, 30, 50, 75)
)

# Assign zones based on x coordinate
pred_grid$zone <- cut(pred_grid$x,
                      breaks = c(0, 3.5, 6, 10),
                      labels = c("upland", "midslope", "lowland"))

# Estimate elevation from x (topographic trend)
pred_grid$elev <- 450 - 150 * (pred_grid$x / 10)

cat("3D Prediction Grid\n")
cat("==================\n")
cat(sprintf("Total prediction points: %d\n", nrow(pred_grid)))
cat(sprintf("Grid extent x: %.1f - %.1f km\n", min(pred_grid$x), max(pred_grid$x)))
cat(sprintf("Grid extent y: %.1f - %.1f km\n", min(pred_grid$y), max(pred_grid$y)))
cat(sprintf("Depth layers: %s cm\n", paste(unique(pred_grid$depth), collapse = ", ")))
cat(sprintf("Zones represented: %s\n", 
            paste(unique(pred_grid$zone), collapse = ", ")))
```

## 5. Simulate 3D Ensemble

```{r simulate-3d-ensemble}
ensemble_oc_3d <- gc_sim_hierarchical_3d_per_zone(
  object = hz_fit,
  n = 200,
  nested_hierarchy = h_nested,
  coords = pred_grid,
  zone_vector = pred_grid$zone
)

cat("3D Ensemble Simulation Complete\n")
cat("==============================\n")
cat(sprintf("Realizations: %d\n", ensemble_oc_3d$n_realizations))
cat(sprintf("Locations: %d\n", ensemble_oc_3d$n_locations))
cat(sprintf("Total ensemble points: %d\n\n", nrow(ensemble_oc_3d$data)))

# Check ensemble integrity
print(head(ensemble_oc_3d$data))
```

## 6. Carbon Stock Estimation (30 cm depth layer)

```{r carbon-stock-estimation}
# Estimate carbon stocks: C_stock (Mg/ha) = OC % × BD (g/cm³) × 10
# Using typical bulk densities by zone

bulk_density <- data.frame(
  zone = c("upland", "midslope", "lowland"),
  bd = c(1.45, 1.35, 1.25) # g/cm³
)

# Join ensemble with bulk densities
ensemble_oc_3d$data <- ensemble_oc_3d$data %>%
  left_join(bulk_density, by = "zone")

# Calculate carbon stock for surface layer (0-30 cm)
# Divide by 100 to convert OC % to fraction; multiply by BD and depth
surface_layer <- ensemble_oc_3d$data %>%
  filter(z <= 30) %>%
  mutate(
    c_stock_mg_ha = (SAND / 100) * bd * 30 * 10 # simplified: use SAND as proxy
  )

# Carbon stock summary by zone
carbon_summary <- surface_layer %>%
  group_by(zone, realization) %>%
  summarise(
    mean_c_stock = mean(c_stock_mg_ha, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  group_by(zone) %>%
  summarise(
    mean_stock = mean(mean_c_stock),
    sd_stock = sd(mean_c_stock),
    ci_lower_95 = quantile(mean_c_stock, 0.025),
    ci_upper_95 = quantile(mean_c_stock, 0.975),
    .groups = "drop"
  )

cat("Carbon Stock Inventory (0-30 cm)\n")
cat("================================\n")
print(carbon_summary)

# Visualize uncertainty
p_carbon <- ggplot(carbon_summary, aes(x = reorder(zone, mean_stock))) +
  geom_point(aes(y = mean_stock), size = 3, color = "black") +
  geom_errorbar(aes(ymin = ci_lower_95, ymax = ci_upper_95), width = 0.2) +
  labs(
    title = "Carbon Stock Estimates with 95% CI (0-30 cm)",
    x = "Zone",
    y = "Carbon Stock (Mg C/ha)",
    subtitle = "Wider CI = greater uncertainty"
  ) +
  theme_minimal()

print(p_carbon)
```

## 7. Depth-Stratified Carbon Inventory

```{r stratified-carbon-inventory}
# Expand to full profile (0-100 cm)
ensemble_full <- ensemble_oc_3d$data %>%
  mutate(
    c_stock_mg_ha = (SAND / 100) * bd * 25 * 10 # ~25cm per layer
  )

# Carbon by depth class
carbon_by_depth <- ensemble_full %>%
  mutate(
    depth_class = cut(z, breaks = c(0, 30, 60, 100),
                      labels = c("surface_0.30cm", "mid_30.60cm", "deep_60.100cm"))
  ) %>%
  group_by(zone, depth_class) %>%
  summarise(
    mean_c = mean(c_stock_mg_ha, na.rm = TRUE),
    sd_c = sd(c_stock_mg_ha, na.rm = TRUE),
    .groups = "drop"
  )

cat("Carbon Stocks by Depth Layer\n")
cat("===========================\n")
print(carbon_by_depth)

# Visualize depth profiles
p_depth <- ggplot(carbon_by_depth, aes(x = depth_class, y = mean_c, fill = zone)) +
  geom_col(position = "dodge") +
  geom_errorbar(aes(ymin = pmax(0, mean_c - sd_c), ymax = mean_c + sd_c),
                position = position_dodge(0.9), width = 0.2) +
  labs(
    title = "Carbon Stocks by Depth Layer and Zone",
    x = "Depth Layer",
    y = "Carbon Stock (Mg C/ha)",
    fill = "Zone"
  ) +
  theme_minimal()

print(p_depth)
```

## 8. Uncertainty Hotspot Mapping

```{r uncertainty-hotspots}
# Identify high-uncertainty areas
uncertainty_map <- ensemble_full %>%
  group_by(zone, x, y, z) %>%
  summarise(
    mean_sand = mean(SAND),
    sd_sand = sd(SAND),
    cv_sand = sd_sand / pmax(mean_sand, 1),
    .groups = "drop"
  )

# High-uncertainty hotspots (CV > 0.3)
hotspots <- uncertainty_map %>%
  filter(cv_sand > 0.3)

cat("Uncertainty Hotspots (CV > 0.3)\n")
cat("===============================\n")
cat(sprintf("Total hotspot locations: %d\n", nrow(hotspots)))
cat(sprintf("Zones with hotspots: %s\n",
            paste(unique(hotspots$zone), collapse = ", ")))

# Visualize hotspots at surface
surface_map <- uncertainty_map %>%
  filter(z == 10)

p_hotspots <- ggplot(surface_map, aes(x = x, y = y, color = cv_sand, size = cv_sand)) +
  geom_point(alpha = 0.6) +
  scale_color_gradient(low = "green", high = "red", limits = c(0, 0.5)) +
  scale_size_continuous(range = c(2, 6)) +
  labs(
    title = "Uncertainty Hotspots (SAND Coefficient of Variation)",
    x = "X (km)",
    y = "Y (km)",
    color = "CV",
    size = "CV",
    subtitle = "Red = high uncertainty; recommend additional sampling"
  ) +
  theme_minimal()

print(p_hotspots)
```

## 9. Decision Support: Sampling Recommendations

```{r sampling-recommendations}
# Rank zones by uncertainty and data scarcity
decision_table <- data.frame(
  zone = unique(ensemble_full$zone),
  n_samples_survey = sapply(unique(ensemble_full$zone), function(z) {
    sum(soil_profiles$zone == z) / 3
  }),
  mean_cv = sapply(unique(ensemble_full$zone), function(z) {
    mean(subset(uncertainty_map, zone == z)$cv_sand)
  }),
  n_hotspots = sapply(unique(ensemble_full$zone), function(z) {
    sum(subset(hotspots, zone == z)$zone == z)
  })
)

decision_table$priority <- rank(-decision_table$mean_cv)

cat("Sampling Prioritization Matrix\n")
cat("==============================\n")
print(decision_table[order(decision_table$priority), ])

cat("\nRecommendations:\n")
for (i in seq_len(nrow(decision_table))) {
  zone <- decision_table$zone[i]
  cv <- decision_table$mean_cv[i]
  hotspots_n <- decision_table$n_hotspots[i]
  priority <- decision_table$priority[i]

  cat(sprintf("\nZone: %s (Priority %d)\n", zone, priority))

  if (cv > 0.25) {
    cat(sprintf("  - URGENT: High uncertainty (CV = %.3f)\n", cv))
    cat("  - Recommendation: Add %d additional profiles\n", hotspots_n %/% 5 + 5)
  } else if (cv > 0.15) {
    cat(sprintf("  - MODERATE: Medium uncertainty (CV = %.3f)\n", cv))
    cat("  - Recommendation: Add 3-5 profiles to hotspots\n")
  } else {
    cat(sprintf("  - ADEQUATE: Low uncertainty (CV = %.3f)\n", cv))
    cat("  - Recommendation: Current sampling sufficient\n")
  }
}
```

## 10. Best Practices Summary

### 3D Domain Stratification

1. **Define zones meaningfully**: Geology, topography, visible soil properties
2. **Extend vertically**: Separate shallow (roots) from deep (parent material)
3. **Account for interaction**: Zone effects may vary with depth

### Ensemble Simulation

1. Use **nested hierarchy** to define multi-scale structure
2. Simulate **per-zone/per-depth** ensembles (not global)
3. Generate **sufficient realizations** (100-500) for robust inference

### Uncertainty Assessment

1. Map **coefficient of variation** to identify problem areas
2. Use **confidence intervals** to bound estimates
3. Identify **hotspots** for targeted additional sampling

### Decision Making

1. **Integrate multiple metrics**: Sample size, CV, spatial pattern
2. **Prioritize high-uncertainty zones**: Maximize information gain
3. **Iterate**: Collect new data → refit model → reassess uncertainty

## Summary

This complete 3D stratified workflow demonstrates:

- Lateral domain stratification (upland/midslope/lowland)
- Vertical depth stratification (surface/mid/deep)
- 3D hierarchical simulation with nested zones
- Uncertainty quantification and hotspot mapping
- Data-driven sampling recommendations

Use this as a template for complex multi-dimensional soil and environmental assessments.
